---
title: "compute_aldex_clr"
author: "Chiamh"
date: '2024-04-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Scripts used to compute centered log ratio transformed abundances using aldex2


```{r}
library(tidyverse)
library(ALDEx2) #v1.28.1
library(propr) #version 2.1.2

```

Load supporting files and inputs
```{r}

metadata <- read_tsv("../metadata/skin_mtx_metadata_fmt.txt", show_col_types = FALSE) 

mgx_stats <- read_tsv("../metadata/MGX_QC_stats.txt", show_col_types = FALSE) 

mtx_stats <- read_tsv("../metadata/MTX_QC_stats.txt", show_col_types = FALSE) 

mgx_stats <- merge(mgx_stats, metadata, by = "LIBID", all.x=TRUE)

mtx_stats <- merge(mtx_stats, metadata, by ="LIBID", all.x=TRUE)


mtx_to_pull <- read_tsv("../metadata/mtx_to_pull", col_names = FALSE, show_col_types = FALSE) %>% pull(X1)

mgx_to_pull <- read_tsv("../metadata/mgx_to_pull", col_names = FALSE, show_col_types = FALSE) %>% pull(X1)

mtx_stats_chosen <- mtx_stats %>% dplyr::filter(LIBID %in% mtx_to_pull)

mgx_stats_chosen <- mgx_stats %>% dplyr::filter(LIBID %in% mgx_to_pull)

mtx_mgx_stats_chosen <- read_tsv(file="../metadata/mtx_mgx_stats_chosen.tsv", show_col_types = FALSE)

##########
#MTX Library IDs per site

mtx_Sc_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Sc") %>% pull(mtx_LIBID)
mtx_Ch_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Ch") %>% pull(mtx_LIBID)
mtx_Ac_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Ac") %>% pull(mtx_LIBID)
mtx_Vf_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Vf") %>% pull(mtx_LIBID)
mtx_Tw_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Tw") %>% pull(mtx_LIBID)

#MGX Library IDs per site

mgx_Sc_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Sc") %>% pull(mgx_LIBID)
mgx_Ch_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Ch") %>% pull(mgx_LIBID)
mgx_Ac_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Ac") %>% pull(mgx_LIBID)
mgx_Vf_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Vf") %>% pull(mgx_LIBID)
mgx_Tw_ids <- mtx_mgx_stats_chosen %>% dplyr::filter(region=="Tw") %>% pull(mgx_LIBID)


#rows are taxa and columns are samples
bracken_count_mat <- read.table(file="../data/bracken/mgx_bracken_readcounts_mat", header = T, comment = '', check = F, sep = ' ')

#convert to dataframe
bracken_count_df <- as.data.frame(bracken_count_mat)


```



The ALDEx2 R package (version 1.18.0)15 generates Monte Carlo samples of Dirichlet distributions for each sample, using a uniform prior, and performs CLR transformation of each realization."

Prevalence filtering should be done PER site, resulting in 5 separate clr transformed bracken abundance tables
Regarding prevalence filtering, Nearing et al. (2022) found that applying a 10% threshold for the prevalence of the taxa generally resulted in more robust results. 

aldex.clr takes a "reads" data.frame with each row being a different gene/microbe and each column represents the read count from that library

Try mc.samples = 1000 is usually sufficient

These functions take awhile to run. 

Pre-run results can be downloaded from figshare:
https://figshare.com/account/projects/202683/articles/25699095

```{r}

#helper function to compute aldex clr values, averaged across all monte carlo instances

get_mean_aldex_clr <- function(counts_in,seed=123){
  
  set.seed(seed)
  
  aldex_clr_results <- aldex.clr(counts_in, mc.samples=1000, verbose = FALSE)
  
  aldex_propr_object <- aldex2propr(aldex_clr_results)
  output <- aldex_propr_object@logratio #contains the log-ratio transformed counts as averaged across all Monte Carlo instances
  
  return(output)
}

#Returning just the raw clr values

get_raw_aldex_clr <- function(counts_in,seed=123){
  
  set.seed(seed)
  
  aldex_clr_results <- aldex.clr(counts_in, mc.samples=1000, verbose = FALSE)
  
  return(aldex_clr_results)
}


#default 10% prevalence
compute_aldex_inputs <- function(df=bracken_count_df, target_prevalence=0.1){
  
  Sc_n_libs <- length(mgx_Sc_ids)
  Ch_n_libs <- length(mgx_Ch_ids)
  Ac_n_libs <- length(mgx_Ac_ids)
  Vf_n_libs <- length(mgx_Vf_ids)
  Tw_n_libs <- length(mgx_Tw_ids)
  
  #For Sc
  Sc_counts <- df %>% dplyr::select(all_of(mgx_Sc_ids))
  Sc_counts <- Sc_counts[rowSums(Sc_counts==0) < (target_prevalence*Sc_n_libs),]
  Sc_bracken_clr <- get_mean_aldex_clr(Sc_counts)
  
  #For Ch
  Ch_counts <- df %>% dplyr::select(all_of(mgx_Ch_ids))
  Ch_counts <- Ch_counts[rowSums(Ch_counts==0) < (target_prevalence*Ch_n_libs),]
  Ch_bracken_clr <- get_mean_aldex_clr(Ch_counts)
  
  #For Ac
  Ac_counts <- df %>% dplyr::select(all_of(mgx_Ac_ids))
  Ac_counts <- Ac_counts[rowSums(Ac_counts==0) < (target_prevalence*Ac_n_libs),]
  Ac_bracken_clr <- get_mean_aldex_clr(Ac_counts)
  
  #For Vf
  Vf_counts <- df %>% dplyr::select(all_of(mgx_Vf_ids))
  Vf_counts <- Vf_counts[rowSums(Vf_counts==0) < (target_prevalence*Vf_n_libs),]
  Vf_bracken_clr <- get_mean_aldex_clr(Vf_counts)
  
  #For Tw
  Tw_counts <- df %>% dplyr::select(all_of(mgx_Tw_ids))
  Tw_counts <- Tw_counts[rowSums(Tw_counts==0) < (target_prevalence*Tw_n_libs),]
  Tw_bracken_clr <- get_mean_aldex_clr(Tw_counts)

  #lst of dataframes
  
  output <- tibble::lst(Sc_bracken_clr, Ch_bracken_clr,
                        Ac_bracken_clr, Vf_bracken_clr,
                        Tw_bracken_clr)
  return(output)
   
}


compute_raw_aldex <- function(df=bracken_count_df, target_prevalence=0.1){
  
  Sc_n_libs <- length(mgx_Sc_ids)
  Ch_n_libs <- length(mgx_Ch_ids)
  Ac_n_libs <- length(mgx_Ac_ids)
  Vf_n_libs <- length(mgx_Vf_ids)
  Tw_n_libs <- length(mgx_Tw_ids)
  
  #For Sc
  Sc_counts <- df %>% dplyr::select(all_of(mgx_Sc_ids))
  Sc_counts <- Sc_counts[rowSums(Sc_counts==0) < (target_prevalence*Sc_n_libs),]
  Sc_bracken_clr <- get_raw_aldex_clr(Sc_counts)
  
  #For Ch
  Ch_counts <- df %>% dplyr::select(all_of(mgx_Ch_ids))
  Ch_counts <- Ch_counts[rowSums(Ch_counts==0) < (target_prevalence*Ch_n_libs),]
  Ch_bracken_clr <- get_raw_aldex_clr(Ch_counts)
  
  #For Ac
  Ac_counts <- df %>% dplyr::select(all_of(mgx_Ac_ids))
  Ac_counts <- Ac_counts[rowSums(Ac_counts==0) < (target_prevalence*Ac_n_libs),]
  Ac_bracken_clr <- get_raw_aldex_clr(Ac_counts)
  
  #For Vf
  Vf_counts <- df %>% dplyr::select(all_of(mgx_Vf_ids))
  Vf_counts <- Vf_counts[rowSums(Vf_counts==0) < (target_prevalence*Vf_n_libs),]
  Vf_bracken_clr <- get_raw_aldex_clr(Vf_counts)
  
  #For Tw
  Tw_counts <- df %>% dplyr::select(all_of(mgx_Tw_ids))
  Tw_counts <- Tw_counts[rowSums(Tw_counts==0) < (target_prevalence*Tw_n_libs),]
  Tw_bracken_clr <- get_raw_aldex_clr(Tw_counts)

  #lst of aldex2clr objects
  
  output <- tibble::lst(Sc_bracken_clr, Ch_bracken_clr,
                        Ac_bracken_clr, Vf_bracken_clr,
                        Tw_bracken_clr)
  return(output)
   
}




aldex_clr_values <- compute_aldex_inputs()


aldex_raw_clr_values <- compute_raw_aldex()



```
