---
title: "HIBIT_ELISA_analysis"
author: "Chiamh"
date: "2024-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggbreak)
library(FSA)
library(ggpubr)
```
Define functions

```{r}

# Function to count white spaces in a single string
count_spaces <- function(text) {
  space_count <- length(gregexpr(" ", text)[[1]])
  if (space_count == 1 && gregexpr(" ", text)[[1]][1] == -1) {
    return(0)  # Handle case where there are no spaces
  }
  return(space_count)
}


```


Load HIBIT and ELISA datasets (A*SRL).

There was one contaminated sample in this dataset: 8-043 S. epi Ch. Contact Stephen Wearne at A*SRL for more details.

```{r}

IL1B_HIBIT <- read_tsv("../data/wet_lab/pro_IL1B_HiBIT_RLU.txt", show_col_types = FALSE)

IL1B_ELISA <- read_tsv("../data/wet_lab/cleaved_IL1B_ELISA.txt", show_col_types = FALSE)


IL1B_HIBIT$temp <- sapply(IL1B_HIBIT$condition, count_spaces)
IL1B_HIBIT$grouping <- ifelse(IL1B_HIBIT$temp == 3, str_extract(string=IL1B_HIBIT$condition,pattern="S. [A-Za-z]*"),
                              IL1B_HIBIT$condition)


IL1B_ELISA$temp <- sapply(IL1B_ELISA$condition, count_spaces)
IL1B_ELISA$grouping <- ifelse(IL1B_ELISA$temp == 3, str_extract(string=IL1B_ELISA$condition,pattern="S. [A-Za-z]*"),
                              IL1B_ELISA$condition)

condition_to_grouping <- IL1B_HIBIT %>% dplyr::select(condition, grouping) %>% unique(.)

# Filter away contaminated sample
IL1B_HIBIT <- IL1B_HIBIT %>% dplyr::filter(condition != "8-043 S. epi Ch")
IL1B_ELISA <- IL1B_ELISA %>% dplyr::filter(condition != "8-043 S. epi Ch")

```

#Plot HIBIT RLU

For HIBIT (taken from the Promega kit manual)

The luminescent kinetic measurements collected from endogenously expressed proteins provide several degradation parameters that can be used to characterize different PROTAC compounds or responses from different family members to the same PROTAC compound. These parameters include the degradation rate, degradation maximum (Dmax ), and DC50 . To generate degradation profiles from which to calculate these parameters, the relative luminescent units (RLUs) for each PROTAC concentration must first be normalized to the replicate averaged No PROTAC condition at every time point to account for changes in free furimazine concentration over time. This results in a degradation curve that is expressed as “Fractional RLU” (Equation 1). 

Equation 1: Fractional RLU =  RLU(PROTAC)/ replicate averaged RLU(no PROTAC)

Fractional RLUs were provided in the input files by Stephen

Two biological repeats, three technical replicates each.

Perform Kruskal-Wallis and post hoc tests

```{r}
#for each biological repeat, get the average fractional RLU of the technical replicates. 

IL1B_HIBIT_avg <- IL1B_HIBIT %>% group_by(batch, condition) %>% 
  summarise(avg_over_tech_rep=mean(RLU)) %>% ungroup()

#barplot of HIBIT RLU distribution per species/experimental condition


IL1B_HIBIT_grouping_avg <- merge(IL1B_HIBIT_avg, condition_to_grouping, by = "condition") 

IL1B_HIBIT_grouping_avg_to_plot <- IL1B_HIBIT_grouping_avg %>% dplyr::filter(!grouping %in% c("untreated","1uM ANS"))

#Perform Kruskal-Wallis and post hoc tests
kruskal.test(avg_over_tech_rep ~ grouping,data=IL1B_HIBIT_grouping_avg_to_plot)

IL1B_HIBIT_dunn_test <- dunnTest(avg_over_tech_rep ~ grouping,data=IL1B_HIBIT_grouping_avg_to_plot, method="bh")

#https://www.datanovia.com/en/blog/ggpubr-how-to-add-p-values-generated-elsewhere-to-a-ggplot/
#format the dunn test results to be compatible with ggpubr::stat_pvalue_manual()

IL1B_HIBIT_dunn_test_df <- IL1B_HIBIT_dunn_test$res

#only selected comparisons for plotting
IL1B_HIBIT_dunn_test_df_subset <- IL1B_HIBIT_dunn_test_df %>%
  dplyr::filter(Comparison %in% c("S. cap - S. epi",
                                  "S. cap - S. hom",
                                  "S. cap - S. war",
                                  "S. epi - S. hom",
                                  "S. epi - S. war"))


IL1B_HIBIT_dunn_test_df_subset <- IL1B_HIBIT_dunn_test_df_subset %>% 
  separate_wider_delim(Comparison, " - ", names=c("group1","group2")) %>% 
  dplyr::rename("p"="P.unadj", "p.adj"="P.adj")

#three significant figures
IL1B_HIBIT_dunn_test_df_subset$p.adj_fmt <- formatC(IL1B_HIBIT_dunn_test_df_subset$p.adj, format = "e", digits = 2)

```
Boxplot of IL1B HIBIT

```{r}
IL1B_HIBIT_plot <- ggplot(IL1B_HIBIT_grouping_avg_to_plot , aes(x=grouping, y=avg_over_tech_rep)) + 
  geom_boxplot() + theme_classic() + 
  ylab("pro-IL-1B Relative Light Units") + 
  theme(axis.title.x = element_blank()) + 
  stat_pvalue_manual(
    IL1B_HIBIT_dunn_test_df_subset, label = "p.adj_fmt",
    y.position = 2.3, step.increase = 0.1) + annotate("text", x=2.5, y = 4, 
                                                      label ="Kruskal-Wallis chi-squared = 19.854, df = 5, p-value = 0.001331")
      

ggsave(plot=IL1B_HIBIT_plot, filename="../plots/IL1B_HIBIT_plot.pdf", height =7, width = 5)

IL1B_HIBIT_plot

```
Repeat the same logic for ELISA results

```{r}

#for each biological repeat, get the average fractional RLU of the technical replicates. 

IL1B_ELISA_avg <- IL1B_ELISA %>% group_by(batch, condition) %>% 
  summarise(avg_over_tech_rep=mean(concentration)) %>% ungroup()

#barplot of IL1B concentrations per species/experimental condition

IL1B_ELISA_grouping_avg <- merge(IL1B_ELISA_avg, condition_to_grouping, by = "condition")

IL1B_ELISA_grouping_avg_to_plot <- IL1B_ELISA_grouping_avg %>% dplyr::filter(!grouping %in% c("untreated","1uM ANS"))

#Perform Kruskal-Wallis and post hoc tests
kruskal.test(avg_over_tech_rep ~ grouping,data=IL1B_ELISA_grouping_avg_to_plot)
#Kruskal-Wallis chi-squared = 33.864, df = 5, p-value = 2.534e-06

IL1B_ELISA_dunn_test <- dunnTest(avg_over_tech_rep ~ grouping,data=IL1B_ELISA_grouping_avg_to_plot, method="bh")

#https://www.datanovia.com/en/blog/ggpubr-how-to-add-p-values-generated-elsewhere-to-a-ggplot/
#format the dunn test results to be compatible with ggpubr::stat_pvalue_manual()

IL1B_ELISA_dunn_test_df <- IL1B_ELISA_dunn_test$res

#only selected comparisons for plotting
IL1B_ELISA_dunn_test_df_subset <- IL1B_ELISA_dunn_test_df %>%
  dplyr::filter(Comparison %in% c("S. cap - S. epi",
                                  "S. cap - S. hom",
                                  "S. cap - S. war",
                                  "S. epi - S. hom",
                                  "S. epi - S. war"))


IL1B_ELISA_dunn_test_df_subset <- IL1B_ELISA_dunn_test_df_subset %>% 
  separate_wider_delim(Comparison, " - ", names=c("group1","group2")) %>% 
  dplyr::rename("p"="P.unadj", "p.adj"="P.adj")

#scientific notation
IL1B_ELISA_dunn_test_df_subset$p.adj_fmt <- formatC(IL1B_ELISA_dunn_test_df_subset$p.adj, format = "e", digits = 2)



```
ELISA boxplot


```{r}
IL1B_ELISA_plot <- ggplot(IL1B_ELISA_grouping_avg %>% dplyr::filter(!grouping %in% c("untreated","1uM ANS")), aes(x=grouping, y=avg_over_tech_rep)) + geom_boxplot() + theme_classic() +
  ylab("cleaved IL-1B concentration") + theme(axis.title.x = element_blank()) +
    stat_pvalue_manual(
    IL1B_ELISA_dunn_test_df_subset, label = "p.adj_fmt",
    y.position = 300, step.increase = 0.1) + annotate("text", x=2.5, y = 525, 
                                                      label ="Kruskal-Wallis chi-squared = 33.864, df = 5, p-value = 2.534e-06")

ggsave(plot=IL1B_ELISA_plot, filename="../plots/IL1B_ELISA_plot.pdf", height =7, width = 5)

IL1B_ELISA_plot
```
