---
title: "human_pangenome_decont"
author: "Chiamh"
date: "2025-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(tidyverse)

```

Load metadata

```{r}

mgx_to_pull <- read_tsv("../metadata/mgx_to_pull", show_col_types = FALSE, col_names = "LIBID")

mgx_stats_chosen <- read_tsv("../metadata/mtx_mgx_stats_chosen.tsv", show_col_types = FALSE) %>% 
  dplyr::select(mgx_LIBID, region, subj_region) %>% dplyr::rename(LIBID=mgx_LIBID)

```


# Motivation

The starting point are WGS reads (from n = 102 samples, full dataset) pre-filtered by bwa mapping to hg38.

All unmapped reads were classified with Kraken2 (containing the human genome in the database) and reads not classified as metazoan (or descendants of) were extracted.

These extracted reads were then mapped again to a human pangenome graph *.gbz (hprc-v1.1-mc-chm13) using giraffe from vg 1.63.1

# Analysis

We can then ask the question: of all reads classified as "non-metazoan" by Kraken2, what percentage mapped to the human pangenome?

```{r}
#This analysis also includes unclassified reads

MGX_hupan_QC <- read_tsv("../data/human_pangenome_check/MGX_microbial_hprc-v1.1-mc-chm13_QC.txt", 
                         col_names = c("LIBID", "non_metazoan_kraken2_read_pairs", "mapped_to_human_pangenome_read_pairs"),
                         show_col_types = FALSE)


MGX_hupan_QC$percentage_mapped <- (MGX_hupan_QC$mapped_to_human_pangenome_read_pairs / MGX_hupan_QC$non_metazoan_kraken2_read_pairs) *100


range(MGX_hupan_QC$percentage_mapped) #0.0007050937 0.3081946092
#median percentage mapped
median(MGX_hupan_QC$percentage_mapped) #0.0137 %



```
median 0.01% is well under the threshold of 0.1% for which species were removed from each metagenome. 

We can further refine this question and ask how many classified, non-metazoan reads were mapped to the human pangenome

```{r}

#all unclassified read pairs in the 102 MGX samples, regardless of mapping to human pangenome 

MGX_all_unclassified <- read_tsv("../data/human_pangenome_check/MGX_unclassified.k2report",
                                 show_col_types = FALSE,
                                 col_names = c("LIBID","unclassified_kraken2_read_pairs","X3")) %>%
  dplyr::select(-X3) %>% dplyr::filter(LIBID %in% MGX_hupan_QC$LIBID)


#This is the # of kraken2 unclassified read pairs from the subset of non-metazoan reads that mapped to the human pangenome

MGX_k2_unclassified_hupan_mapped <- read_tsv("../data/human_pangenome_check/mapped_to_hprc/MGX_k2-unclassified_hupan-mapped.tsv",
                                             show_col_types = FALSE,
                                             col_names = c("LIBID", "unclassified_kraken2_hupan_mapped_read_pairs")) %>%
  dplyr::select(-X3)

#can also use purrr::reduce
MGX_unclassified_df <- merge(MGX_all_unclassified, MGX_k2_unclassified_hupan_mapped, by = "LIBID")

MGX_hupan_QC_fmt <- merge(MGX_hupan_QC, MGX_unclassified_df, by = "LIBID" )

MGX_hupan_QC_fmt$classified_kraken2_read_pairs <- MGX_hupan_QC_fmt$non_metazoan_kraken2_read_pairs - 
  MGX_hupan_QC_fmt$unclassified_kraken2_read_pairs

MGX_hupan_QC_fmt$classified_kraken2_hupan_mapped_read_pairs <- MGX_hupan_QC_fmt$mapped_to_human_pangenome_read_pairs -
  MGX_hupan_QC_fmt$unclassified_kraken2_hupan_mapped_read_pairs

MGX_hupan_QC_fmt$k2_classified_percentage_mapped <- (MGX_hupan_QC_fmt$classified_kraken2_hupan_mapped_read_pairs / MGX_hupan_QC_fmt$classified_kraken2_read_pairs ) *100

range(MGX_hupan_QC_fmt$k2_classified_percentage_mapped) # 0.0006510942 0.1712844554

median(MGX_hupan_QC_fmt$k2_classified_percentage_mapped) #0.008317539


```
We can ask another question: across all 5 body sites, what % of different microbial genus level reads (sans contaminants) are mapped to human pangenome?

The genera to check are: Cutibacterium, Staphylococcus and Malassezia

```{r}

#Load genus level kraken2 results for read pairs that were mapped to the human pangenome (hupan)

mgx_k2_genus_hupan_mapped <- lapply(mgx_to_pull$LIBID, function(ID){
  
  df <- read_tsv(paste0("../data/human_pangenome_check/mapped_to_hprc/",ID,"_microbe_to_human_genus.kreport2"),
                 show_col_types = FALSE,
                 col_names = c("read_pairs_mapped_to_hupan", "genus")) %>%
    dplyr::filter(genus %in% c("Cutibacterium","Staphylococcus","Malassezia"))
  
  df$LIBID <- ID
  
  return(df)
  
}) %>% do.call("rbind",.)


#Load genus level kraken2 results for all MGX read pairs, including those not mapped to human pangenome

mgx_k2_genus <- lapply(mgx_to_pull$LIBID, function(ID){
  
  df <- read_tsv(paste0("../data/kraken2/DNA/",ID,"_merged_k2.g.tsv"),
                 show_col_types = FALSE,
                 col_names = c("read_pairs_rooted_at_genus", "genus")) %>%
    dplyr::filter(genus %in% c("Cutibacterium","Staphylococcus","Malassezia"))
  
  df$LIBID <- ID
  
  return(df)
  
}) %>% do.call("rbind",.)

#merge the dataframes

mgx_k2_genus_fmt <- merge(mgx_k2_genus, mgx_k2_genus_hupan_mapped, by = c("LIBID","genus"), all.x=TRUE) %>%
  mutate(across(everything(), ~ replace_na(.x, FALSE)))



mgx_k2_genus_fmt$percentage_mapped_to_hupan <- (mgx_k2_genus_fmt$read_pairs_mapped_to_hupan / mgx_k2_genus_fmt$read_pairs_rooted_at_genus) *100

#median %s
mgx_k2_genus_median <- mgx_k2_genus_fmt %>% group_by(genus) %>% summarize(median=median(percentage_mapped_to_hupan))

# Cutibacterium  0.00108
# Malassezia     0.0258 
# Staphylococcus 0.00304


#Plot these percentages as barplots

mgx_k2_genus_hupan_barplots <- ggplot(mgx_k2_genus_fmt, 
       aes(x=genus, y=percentage_mapped_to_hupan, fill=genus)) + geom_boxplot() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x=element_blank()) + 
  theme(axis.text.x = element_text(face = "italic")) + labs(y = "% mapped to human pangenome")


ggsave(filename="../plots/mgx_k2_genus_hupan_barplots.pdf",
       plot=mgx_k2_genus_hupan_barplots ,
       height = 8, width = 8)


mgx_k2_genus_hupan_barplots 



```








