---
title: "mtx_vs_mgx_counts_species_level"
author: "Chiamh"
date: '2024-04-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Comparison of metatranscriptomic and metagenomic counts for various species across skin sites. 


 
Load libraries

```{r}

library(tidyverse)
library(ggpubr)

source("custom_functions.R")

```

Load metadata and supporting files.


```{r}

#for mgx and mtx data processed using the regular steps of the pipeline described in https://github.com/Chiamh/meta-omics-nf/tree/main
combined_taxa_count_activity_df <- read_tsv("../data/combined_taxa_count_activity_df.tsv",
                                            show_col_types = FALSE)

#for mgx data processed using fastp, STAR, rRNA removal and de-duplication just like mtx reads
mgx_alt_process_taxa_count_activity_df <- read_tsv("../data/mgx_alt_process_taxa_count_activity_df.tsv",
                                                   show_col_types = FALSE) %>%
                                          dplyr::rename(mgx_alt_counts=mgx_counts,
                                                        mgx_alt_RPK=mgx_RPK,
                                                        mgx_alt_CPM=mgx_CPM,
                                                        mgx_alt_CPM_prok=mgx_CPM_prok)


mgx_species_site_prevalence_labelled <- read_tsv("../data/mgx_species_site_prevalence_labelled.tsv",
                                                 show_col_types = FALSE)


#For plotting the species that are at least 50% prevalent (by DNA) per site.
mgx_species_site_prevalence_labelled_threshold <- mgx_species_site_prevalence_labelled %>%
                                                  dplyr::filter(category_within_site!="variable")


```

## Scatter plot

Scatter plots of species level metatranscriptomic (RNA) counts expressed as transcripts per million (TPM)
against species level metagenomic (DNA) counts expressed as counts per million (CPM). Points on the solid line represent
equal proportions of RNA to DNA for a species. The areas beyond the dotted lines represents ≥4-fold differences between
RNA and DNA counts. Plots represent data from all 102 paired metagenomes and metatranscriptomes in this study. Note that
Staphylococcus (dark blue) and Malassezia (lilac) species have an outsized contribution to metatranscriptomes at most sites despite their lower representation in metagenomes.

Data is from the full cohort (n=24, 23, 19, 18, 18 for scalp, cheek, antecubital fossa, volar forearm and toe web respectively).


```{r}
core_prevalent_combined_taxa_count_activity_df <- lapply(c("Sc","Ch","Ac","Vf","Tw"), function(x){
  
  #This will only retain species which are prevalent or core to a site.
  mgx_species_specific_site <- mgx_species_site_prevalence_labelled_threshold %>% 
    dplyr::filter(region==x)
  
  #dataframe with transcriptional activities of prevalent or core species at each site
  specific_site_species_activity_df <- combined_taxa_count_activity_df %>% 
    dplyr::filter(region==x & k2_taxon %in% mgx_species_specific_site$species)
  
  return(specific_site_species_activity_df)
  
}) %>% do.call("rbind",.)




core_prevalent_combined_taxa_count_activity_df$region <- factor(core_prevalent_combined_taxa_count_activity_df$region, 
                                           levels=c("Sc","Ch","Ac","Vf","Tw"))



core_prevalent_combined_taxa_count_activity_df$genus <- word(string=core_prevalent_combined_taxa_count_activity_df$k2_taxon, 
                                                             start=1, sep="_")


core_prevalent_combined_taxa_TPM_scatter <- ggplot(core_prevalent_combined_taxa_count_activity_df, 
       aes(y=mtx_TPM, x=mgx_CPM, color=genus)) + geom_point(size=2) + 
  theme_classic() + scale_color_manual(values=c24) +
  facet_wrap(~region, ncol=3, nrow=2, scales='free') +
  scale_y_continuous(labels = function(x) format(x, scientific = TRUE)) +
  scale_x_continuous(labels = function(x) format(x, scientific = TRUE)) +
  theme(axis.line=element_line())  +
  geom_abline(intercept=0, slope=1) + 
  geom_abline(intercept=0, slope=4,linetype=3) +
  geom_abline(intercept=0, slope=0.25,linetype=3)  
  

ggsave(plot = core_prevalent_combined_taxa_TPM_scatter, 
       filename = "../plots/core_common_taxa_RNA_v_DNA.pdf", width = 10, height = 7)


core_prevalent_combined_taxa_TPM_scatter


```

Repeat the same analysis with mgx CPM values derived from the same mgx input but processed using fastp, STAR, rRNA removal and de-duplication just like mtx reads


```{r}

ALT_combined_taxa_count_activity_df <- merge(combined_taxa_count_activity_df, mgx_alt_process_taxa_count_activity_df,
                                             by=c("mgx_LIBID","k2_taxon"), all.x=TRUE)

ALT_combined_taxa_count_activity_df[is.na(ALT_combined_taxa_count_activity_df)] <- 0


ALT_core_prevalent_combined_taxa_count_activity_df <- lapply(c("Sc","Ch","Ac","Vf","Tw"), function(x){
  
  #This will only retain species which are prevalent or core to a site.
  mgx_species_specific_site <- mgx_species_site_prevalence_labelled_threshold %>% 
    dplyr::filter(region==x)
  
  #dataframe with transcriptional activities of prevalent or core species at each site
  ALT_specific_site_species_activity_df <- ALT_combined_taxa_count_activity_df %>% 
    dplyr::filter(region==x & k2_taxon %in% mgx_species_specific_site$species)
  
  return(ALT_specific_site_species_activity_df)
  
}) %>% do.call("rbind",.)


ALT_core_prevalent_combined_taxa_count_activity_df$region <- factor(ALT_core_prevalent_combined_taxa_count_activity_df$region, 
                                           levels=c("Sc","Ch","Ac","Vf","Tw"))



ALT_core_prevalent_combined_taxa_count_activity_df$genus <- word(string=ALT_core_prevalent_combined_taxa_count_activity_df$k2_taxon, 
                                                             start=1, sep="_")

ALT_core_prevalent_combined_taxa_TPM_scatter <- ggplot(ALT_core_prevalent_combined_taxa_count_activity_df, 
       aes(y=mtx_TPM, x=mgx_alt_CPM, color=genus)) + geom_point(size=2) + 
  theme_classic() + scale_color_manual(values=c24) +
  facet_wrap(~region, ncol=3, nrow=2, scales='free') +
  scale_y_continuous(labels = function(x) format(x, scientific = TRUE)) +
  scale_x_continuous(labels = function(x) format(x, scientific = TRUE)) +
  theme(axis.line=element_line())  +
  geom_abline(intercept=0, slope=1) + 
  geom_abline(intercept=0, slope=4,linetype=3) +
  geom_abline(intercept=0, slope=0.25,linetype=3)  
  

ggsave(plot = ALT_core_prevalent_combined_taxa_TPM_scatter, 
       filename = "../plots/ALT_core_common_taxa_RNA_v_DNA.pdf", width = 10, height = 7)


ALT_core_prevalent_combined_taxa_TPM_scatter


```

Correlate mgx rel abun and mgx rel abun at genus level.

```{r}

#Color code by genus

mgx_alt_vs_mgx_rel_abun_corr <- ggplot(data=ALT_core_prevalent_combined_taxa_count_activity_df, aes(x=mgx_rel_abun,
                                                      y=mgx_alt_rel_abun,
                                                      )) + 
  geom_point(aes(color=genus, size=5)) + theme_classic() + stat_cor(method = "pearson") + 
  stat_smooth( method = "lm") + scale_color_manual(values=c24)


ggsave("../plots/mgx_alt_vs_mgx_rel_abun_corr.pdf", plot=mgx_alt_vs_mgx_rel_abun_corr)

mgx_alt_vs_mgx_rel_abun_corr


```






