---
title: "species_level_DE"
author: "Chiamh"
date: "2024-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}

library(tidyverse)
library(DESeq2)
library(RColorBrewer)
library(pheatmap)

library("clusterProfiler") #v4.4.4
library("GSEABase")
library(EnhancedVolcano)
library(limma)

library(ggbreak)

library(tximport)

```

Load metadata

```{r}

metadata <- read_tsv("../metadata/skin_mtx_metadata_fmt.txt", show_col_types = FALSE) 

#The library IDs of the 102 mtx libraries in this project
mtx_to_pull <- read_tsv("../metadata/mtx_to_pull", col_names = "LIBID") %>% pull(LIBID)

mtx_stats <- read_tsv("../metadata/MTX_QC_stats.txt", show_col_types = FALSE) 

mtx_stats <- merge(mtx_stats, metadata, by ="LIBID", all.x=TRUE)

mtx_stats_chosen <- mtx_stats %>% dplyr::filter(LIBID %in% mtx_to_pull)


mtx_mgx_stats_chosen <- read_tsv(file="../metadata/mtx_mgx_stats_chosen.tsv",
                                show_col_types = FALSE)


S_epi_in_vitro_metadata <- read_tsv("../metadata/S_epi_in_vitro.txt")



#library names of in house S epidermidis culture experiments
#OS: osmotic stress
#SP: stationary phase

in_house_S_epi_in_vitro <- c("Sepi_ctrl1",
                             "Sepi_ctrl2",
                             "Sepi_ctrl3",
                             "Sepi_OS1",
                             "Sepi_OS2",
                             "Sepi_OS3",
                             "Sepi_SP1",
                             "Sepi_SP2",
                             "Sepi_SP3")



#Beware of commas in the ECnum column.. need to split them into separate rows
#Use the EC annotations from the original propan annotation files instead of from just the emapper2.1.6 output
#S epidermidis pangene annotations
S_epidermidis_anno <- read_csv("../metadata/1282.Staphylococcus.epidermidis.pangenome.annotation.matrix.csv") %>% dplyr::rename(propan_clusterID=Clusters)

S_epidermidis_pangene_names <- read_tsv("../metadata/S_epidermidis_derep_pangenome_names.txt", col_names = "pangene")

#Extract the ECnums from S_epidermidis_anno 

S_epi_propan_EC <- S_epidermidis_anno %>% dplyr::select(propan_clusterID,`Enzyme Class`)
colnames(S_epi_propan_EC)[2] <- "propan_EC_desc"

#separate rows by ";"
S_epi_propan_EC <- S_epi_propan_EC %>% dplyr::filter(propan_EC_desc != "-") %>% 
  separate_rows(propan_EC_desc, sep="; ")


#This regex utilizes lookahead and lookbehind to return the ECNum
#CAUTION: if I did not separate the rows by ";", then I would have needed to use str_extract_all instead, since str_extract only gives the first match
S_epi_propan_EC$ECnum <- str_extract(S_epi_propan_EC$propan_EC_desc, pattern="(?<=\\[)[0-9\\.]*(?=\\])")

S_epi_propan_EC$propan_EC_desc <- gsub(pattern="\\[[0-9]*\\.[0-9]*\\.[0-9]*\\.[0-9]*\\]",replacement="",x=S_epi_propan_EC$propan_EC_desc)




#Load KEGG annotations

KEGG_term2name <- read_tsv("../metadata/KEGG_term2name.tsv", show_col_types = FALSE)

KEGG_hum_disease_pathways <- read_tsv("../metadata/KEGG_hum_disease_pathway_modules.txt", 
                                      show_col_types = FALSE,
                                      col_names = "map_ID")

KEGG_org_sys_pathways <- read_tsv("../metadata/KEGG_organismal_systems_pathway_modules.txt", 
                                  show_col_types = FALSE,
                                  col_names = "map_ID")



#Load GO annotations

# Load GO annotations
GO_annot_all <- read_tsv("../metadata/all_GO_annotations.tsv", show_col_types = FALSE)


#Select annotations associated with "Biological Process"
GO_BP_term2name <- GO_annot_all %>% 
  dplyr::filter(ontology=="BP") %>% 
  dplyr::select(GO_ID, GO_desc) %>%
  dplyr::rename(from=GO_ID, to=GO_desc)


#Select annotations associated with "Molecular Function"
GO_MF_term2name <- GO_annot_all %>% 
  dplyr::filter(ontology=="MF") %>% 
  dplyr::select(GO_ID, GO_desc) %>%
  dplyr::rename(from=GO_ID, to=GO_desc)

#
## Load goslim_metagenome annotations

goslim_metagenome_term2name <- read_tsv("../metadata/goslim_metagenome.txt", 
                                    col_names = c("GO_slim","GO_desc"), show_col_types = FALSE) %>% 
  dplyr::rename(from=GO_slim, to=GO_desc)



#Load eggnog mapper annotations (with GO annotations) for the eight skin bacteria

eight_skin_bact_proteins_emapper_annot <- read_tsv("../metadata/eight_skin_bact_proteins.emapper.annotations",
                                                   comment="#", show_col_types = FALSE,
                                                   col_names = c("query",  
                                                                 "seed_ortholog",   
                                                              "evalue",
                                                              "score",
                                                              "eggNOG_OGs",
                                                              "max_annot_lvl",
                                                              "COG_category",
                                                              "Description",
                                                              "Preferred_name", 
                                                              "GOs", "EC", "KEGG_ko",
                                                              "KEGG_Pathway", "KEGG_Module",
                                                              "KEGG_Reaction",  "KEGG_rclass",
                                                              "BRITE", "KEGG_TC", 
                                                              "CAZy", "BiGG_Reaction", "PFAMs"))



eight_skin_bact_proteins_emapper_annot$species <- word(eight_skin_bact_proteins_emapper_annot$query,1,2,sep="_")

eight_skin_bact_proteins_emapper_annot$propan_clusterID <- word(eight_skin_bact_proteins_emapper_annot$query,3,4,sep="_")



# KO pathway annotations
#https://www.genome.jp/kegg/mapper/search.html
#https://www.genome.jp/kegg-bin/show_organism?menu_type=pathway_maps&org=ko

#ko.pathways <- download_KEGG(species="ko")
#term2name <- ko.pathways$KEGGPATHID2NAME
#write_tsv(term2name, file="../metadata/KEGG_term2name.tsv")

#Load in term2name so that we do not have to download_KEGG every time we re-run this chunk
KEGG_term2name <- read_tsv("../metadata/KEGG_term2name.tsv", show_col_types = FALSE)


```

Define custom functions

```{r}

source("custom_functions.R")

source("DESeq2_wrappers.R")

species_threshold_fn <- function(species, paired_read_threshold, k2_input){
  
  df_filt <- k2_input %>% dplyr::filter(name == species & paired_reads >= paired_read_threshold)
  
  return(df_filt)
  
}


get_lib_vec_fn <- function(in_vivo_site="Tw",in_vitro_condition, in_vitro_study){
  
  #in_vivo IDs
  in_vivo <- DESeq_metadata %>% dplyr::filter(condition %in% in_vivo_site) %>% rownames()
  
  
  #in_vitro IDs
  in_vitro <- S_epi_in_vitro_metadata %>% dplyr::filter(study %in% in_vitro_study & expt_desc %in% in_vitro_condition) %>% pull(Library_ID) %>% unique()
  
  output_vec <- c(in_vivo,in_vitro)
  return(output_vec)
  
}

#Modify the DESeq2 PCA function

plotPCA_custom.DESeqTransform = function(object, intgroup="condition", ntop=500, returnData=FALSE)
{
  # calculate the variance for each gene
  rv <- rowVars(assay(object))

  # select the ntop genes by variance
  select <- order(rv, decreasing=TRUE)[seq_len(min(ntop, length(rv)))]

  # perform a PCA on the data in assay(x) for the selected genes
  pca <- prcomp(t(assay(object)[select,]))

  # the contribution to the total variance for each component
  percentVar <- pca$sdev^2 / sum( pca$sdev^2 )

  if (!all(intgroup %in% names(colData(object)))) {
    stop("the argument 'intgroup' should specify columns of colData(dds)")
  }

  intgroup.df <- as.data.frame(colData(object)[, intgroup, drop=FALSE])
  
  # add the intgroup factors together to create a new grouping factor
  group <- if (length(intgroup) > 1) {
    factor(apply( intgroup.df, 1, paste, collapse=":"))
  } else {
    colData(object)[[intgroup]]
  }

  # assemble the data for the plot
  d <- data.frame(PC1=pca$x[,1], PC2=pca$x[,2],
                  PC3=pca$x[,3], PC4=pca$x[,4],
                  PC5=pca$x[,5], PC6=pca$x[,6],
                  PC7=pca$x[,7], PC8=pca$x[,8],
                  group=group, intgroup.df, name=colnames(object))

  if (returnData) {
    variance_vector <- percentVar
    attr(d, "percentVar") <- percentVar[1:2]
    return(list(d, variance_vector))
  }
  
  ggplot(data=d, aes_string(x="PC1", y="PC2", color="group")) + geom_point(size=5) + 
    xlab(paste0("PC1: ",round(percentVar[1] * 100),"% variance")) +
      ylab(paste0("PC2: ",round(percentVar[2] * 100),"% variance")) +
        coord_fixed() + theme_classic() + scale_color_manual(values=c24)
}

setMethod("plotPCA", signature(object="DESeqTransform"), plotPCA_custom.DESeqTransform)




```

# Malassezia restricta analysis

## Load Malassezia restricta gene annotations

Gene counts from Salmon

```{r}


###Load Malassezia transcriptome eggNOG annotations. Also need to associate each query with a Malassezia locus tag
#This can be done with Malassezia_tx2gene

Malassezia_OG_annot <- read_tsv("../data/emapper_out/Malassezia_CDS.emapper.annotations",
                                                   comment="#", show_col_types = FALSE,
                                                   col_names = c("query",  
                                                                 "seed_ortholog",   
                                                              "evalue",
                                                              "score",
                                                              "eggNOG_OGs",
                                                              "max_annot_lvl",
                                                              "COG_category",
                                                              "Description",
                                                              "Preferred_name", 
                                                              "GOs", "EC", "KEGG_ko",
                                                              "KEGG_Pathway", "KEGG_Module",
                                                              "KEGG_Reaction",  "KEGG_rclass",
                                                              "BRITE", "KEGG_TC", 
                                                              "CAZy", "BiGG_Reaction", "PFAMs"))



#For salmon tximport
Malassezia_tx2gene <- read_tsv("../metadata/Malassezia_tx2gene.tsv", col_names = FALSE, show_col_types = FALSE) #21037 transcripts and 21025 genes

#Add Malassezia gene annotations to the eggnog annotation file
colnames(Malassezia_tx2gene) <- c("query","gene")
Malassezia_OG_annot <- merge(Malassezia_OG_annot, Malassezia_tx2gene, all.x=TRUE, by = "query")


MGL_OG_annot <- Malassezia_OG_annot %>% dplyr::filter(str_detect(string=Malassezia_OG_annot$gene, pattern="MGL"))
MGL_OG_annot$annot <- paste0(MGL_OG_annot$gene,"_",MGL_OG_annot$Description)

#Details of M globosa genes and the 13 secreted lipases
MGL_secretory_lipase <- MGL_OG_annot %>% dplyr::filter(str_detect(string=MGL_OG_annot$Description, pattern="Secretory lipase|^Lipase"))

#
Malassezia_salmon_files <- file.path("../data/mtx_species_mapping/Malassezia_salmon_out", paste0(mtx_to_pull, "_quant.sf"))

names(Malassezia_salmon_files) <- mtx_to_pull

Malassezia_txi.salmon <- tximport(Malassezia_salmon_files, type="salmon", tx2gene=Malassezia_tx2gene)

Malassezia_count_matrix <- Malassezia_txi.salmon$counts

#Convert to matrix of integers
  
Malassezia_integer_count_matrix <- round(Malassezia_count_matrix)
mode(Malassezia_integer_count_matrix) <- "integer"
#write.table(Malassezia_integer_count_matrix, file="../data/Malassezia_integer_count_matrix.txt", row.names=TRUE, col.names=TRUE)

Malassezia_TPM_matrix <- Malassezia_txi.salmon$abundance
#write.table(Malassezia_TPM_matrix, file="../data/Malassezia_TPM_matrix.txt", row.names=TRUE, col.names=TRUE)

#We need to generate a custom term2gene dataframe specific to M restricta 

M_res_term2gene <- get_Malassezia_term2gene_df(species="Malassezia_restricta")
M_res_term2gene <- M_res_term2gene %>% dplyr::filter(!from %in% c(KEGG_hum_disease_pathways$map_ID,
                                               KEGG_org_sys_pathways$map_ID))

MGL_term2gene <- get_Malassezia_term2gene_df(species="Malassezia_globosa")
MGL_term2gene <- MGL_term2gene %>% dplyr::filter(!from %in% c(KEGG_hum_disease_pathways$map_ID,
                                               KEGG_org_sys_pathways$map_ID))





###########
##M restricta
##########
#Malassezia restricta has good annotations from the gtf file
M_res_gtf_annots <- read_tsv("../metadata/M_restricta_GCA_003691605.1_ASM369160v1_cds_from_genomic_annots.tsv", 
                             col_names = c("gene_tag","gene","proteinID"), show_col_types = FALSE)

M_res_gtf_annots$annot <- ifelse(is.na(M_res_gtf_annots$gene_tag) & is.na(M_res_gtf_annots$proteinID), 
                                 M_res_gtf_annots$gene, ifelse(is.na(M_res_gtf_annots$gene_tag)
                                                               & !is.na(M_res_gtf_annots$proteinID), 
                                                               M_res_gtf_annots$proteinID, M_res_gtf_annots$gene_tag)
                                 )
M_res_gtf_annots$annot <- paste0(M_res_gtf_annots$gene,"_",M_res_gtf_annots$annot)
#Shorten very long label
M_res_gtf_annots$annot <- gsub(pattern="putative CDP-alcohol phosphatidyltransferase class-I family protein C22A12.10", 
                               replacement="CDP-alcohol phosphatidyltransferase",
                               M_res_gtf_annots$annot)

```
## Differential expression analysis for Malassezia restricta

```{r}

mtx_Sc_Ch_libs <- mtx_stats_chosen %>% dplyr::filter(region %in% c("Sc","Ch")) %>% pull(LIBID)

M_res_Sc_Ch_count_matrix <- Malassezia_count_matrix[grep("DNF",row.names(Malassezia_count_matrix)),
                                                    mtx_Sc_Ch_libs]



#select those with a minimum M restricta read threshold by Kraken2.
#MTX data

rna_k2_minimizer <- lapply(mtx_to_pull, function(x){
  
  df_out <- read_tsv(file=paste0("../data/k2_species_renorm/RNA/unfiltered/",x,"_k2_minimizer.s.tsv"),
           show_col_types = FALSE)
  return(df_out)
  
})

names(rna_k2_minimizer) <- mtx_to_pull



M_restricta_Sc_Ch <- lapply(rna_k2_minimizer[mtx_Sc_Ch_libs], function(df){
  
  output <- df %>% dplyr::filter(k2_taxon == "Malassezia_restricta") %>% dplyr::rename(LIBID=mtx_LIBID)
  
  output <- merge(output, mtx_stats_chosen %>% dplyr::select(c("LIBID","subj_region", "subject","region","low_conc")),
                  by="LIBID")
  
  return(output)
  
}) %>% do.call("rbind",.)



#Run DESeq2 on samples with at least 200K M restricta read pairs
M_restricta_Sc_Ch_filt_DESeq <- run_DESeq_with_salmon_matrix(input_matrix = M_res_Sc_Ch_count_matrix,
                                                           lib_vec=M_restricta_Sc_Ch %>% 
                                                             dplyr::filter(paired_counts >= 200000) %>%  pull(LIBID),
                                                           cond="Sc",
                                                           ref_cond="Ch",
                                                           run_test = TRUE,
                                                           species="Malassezia_restricta")

plotDispEsts(M_restricta_Sc_Ch_filt_DESeq[["dds"]])

DESeq2::plotMA(M_restricta_Sc_Ch_filt_DESeq[["DE_results_shrunken"]])


summary(M_restricta_Sc_Ch_filt_DESeq[["DE_results_shrunken"]])

M_restricta_Sc_Ch_DE_res <- M_restricta_Sc_Ch_filt_DESeq[["DE_results_shrunken_df"]] %>% dplyr::rename(feature=gene)
```

## Panel A

```{r}

#Volcano plot

###Color the points that are involved in peroxisome pathway, enriched lipid pathways or both 

#points that correspond to "peroxisome" pathway - map04146
M_res_term2gene_peroxisome <- M_res_term2gene %>% dplyr::filter(from == "map04146")
#points that correspond to enriched lipid metabolic pathways
#map00565: Ether lipid metabolism
#map00564: Glycerophospholipid metabolism
#map00600: Sphingolipid metabolism

M_res_term2gene_enriched_lipid_metab <- M_res_term2gene %>% 
  dplyr::filter(from %in%  c("map00561","map00564","map00600"))

M_res_Sc_Ch_volcano_keyvals_color <- ifelse(M_restricta_Sc_Ch_DE_res$feature %in% 
                                              intersect(M_res_term2gene_peroxisome$to,M_res_term2gene_enriched_lipid_metab$to), 'green',
                                            ifelse(M_restricta_Sc_Ch_DE_res$feature %in% M_res_term2gene_peroxisome$to,'royalblue', 
                                                   ifelse(M_restricta_Sc_Ch_DE_res$feature %in% M_res_term2gene_enriched_lipid_metab$to, "gold",
                                            'black')))

M_res_Sc_Ch_volcano_keyvals_color[is.na(M_res_Sc_Ch_volcano_keyvals_color)] <- 'black'
names(M_res_Sc_Ch_volcano_keyvals_color)[M_res_Sc_Ch_volcano_keyvals_color == 'royalblue'] <- 'Peroxisome'
names(M_res_Sc_Ch_volcano_keyvals_color)[M_res_Sc_Ch_volcano_keyvals_color == 'gold'] <- 'Ether lipid/Glycerophospholipid/Sphingolipid metabolism'
names(M_res_Sc_Ch_volcano_keyvals_color)[M_res_Sc_Ch_volcano_keyvals_color == 'green'] <- 'Both'


M_restricta_volcano_gene_labels <- data.frame(gene=rownames(M_restricta_Sc_Ch_DE_res))

M_restricta_volcano_gene_labels <- merge(M_restricta_volcano_gene_labels, M_res_gtf_annots, all.x=TRUE, by ="gene")

M_restricta_volcano_gene_labels$gene_annot <- ifelse(is.na(M_restricta_volcano_gene_labels$gene_tag),
                                                     M_restricta_volcano_gene_labels$gene,
                                                     M_restricta_volcano_gene_labels$gene_tag)


M_restricta_Sc_Ch_DE_volcano <- EnhancedVolcano(M_restricta_Sc_Ch_DE_res,
                lab = M_restricta_volcano_gene_labels$gene_annot,
                title = "Malassezia restricta transcripts, Sc vs Ch",
                x='log2FoldChange',
                y='padj',
                selectLab = c('plcN_1', 'plcN_2',
                              'plcN_3', 'plcN_4',
                              'plcN_5','plcN_6',
                              'plcN_7', 'stt4', 'pik1',
                              'Acot8','POT1', 'LACS7', 'POX2', 'PEX11B'),
                #boxedLabels = TRUE,
                drawConnectors = TRUE,
                labFace = 'bold',
                labSize = 3.0,
                pCutoff = 0.05,
                FCcutoff=0,
                colCustom = M_res_Sc_Ch_volcano_keyvals_color)

ggsave(filename="../plots/M_restricta_Sc_Ch_DE_volcano.pdf",
       plot=M_restricta_Sc_Ch_DE_volcano,
       height = 8, width = 10)


M_restricta_Sc_Ch_DE_volcano

```
## Panel B

Gene set enrichment analysis for M restricta results

```{r}

##GSEA with KEGG

M_restricta_Sc_Ch_DE_GSEA_KEGG <- GSEA_from_DESeq(genelist = M_restricta_Sc_Ch_filt_DESeq$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = M_restricta_Sc_Ch_filt_DESeq$DE_results_shrunken_df$gene,
                                         term2gene = M_res_term2gene,
                                         term2name = KEGG_term2name)

M_restricta_Sc_Ch_DE_NES_KEGG <- make_NES_barplot(M_restricta_Sc_Ch_DE_GSEA_KEGG@result, title= "Malassezia restricta transcripts, Sc vs Ch")

ggsave(filename="../plots/M_restricta_Sc_Ch_DE_NES_KEGG.pdf",
       plot=M_restricta_Sc_Ch_DE_NES_KEGG,
       height = 8, width = 10)


M_restricta_Sc_Ch_DE_NES_KEGG

```

# Malassezia globosa analysis

## Malassezia globosa comparisons between Scalp and cheek

```{r}

M_globosa_Sc_Ch_count_matrix <- Malassezia_count_matrix[grep("MGL",row.names(Malassezia_count_matrix)),
                                                    mtx_Sc_Ch_libs]


M_globosa_Sc_Ch <- lapply(rna_k2_minimizer[mtx_Sc_Ch_libs], function(df){
  
  output <- df %>% dplyr::filter(k2_taxon == "Malassezia_globosa") %>% dplyr::rename(LIBID=mtx_LIBID)
  
  output <- merge(output, mtx_stats_chosen %>% dplyr::select(c("LIBID","subj_region", "subject","region","low_conc")),
                  by="LIBID")
  
  return(output)
  
}) %>% do.call("rbind",.)



M_globosa_Sc_Ch_PCA_input <- run_DESeq_with_salmon_matrix(input_matrix = M_globosa_Sc_Ch_count_matrix,
                                                           lib_vec=M_globosa_Sc_Ch %>% dplyr::filter(paired_counts >= 200000) %>% pull(LIBID),
                                                           cond="Sc",
                                                           ref_cond="Ch",
                                                           run_test = FALSE,
                                                           species="Malassezia_globosa")

plotPCA(M_globosa_Sc_Ch_PCA_input[["vsd"]], intgroup = "region", returnData=FALSE) + geom_text_repel(aes(label = name))
```
Run DESeq2 for M globosa between scalp and cheeks.

```{r}

M_globosa_Sc_Ch_filt_DESeq <- run_DESeq_with_salmon_matrix(input_matrix = M_globosa_Sc_Ch_count_matrix,
                                                           lib_vec=M_globosa_Sc_Ch %>% 
                                                             dplyr::filter(paired_counts >= 200000) %>%  pull(LIBID),
                                                           cond="Sc",
                                                           ref_cond="Ch",
                                                           run_test = TRUE,
                                                           species="Malassezia_globosa")

plotDispEsts(M_globosa_Sc_Ch_filt_DESeq[["dds"]])

DESeq2::plotMA(M_globosa_Sc_Ch_filt_DESeq[["DE_results_shrunken"]])


summary(M_globosa_Sc_Ch_filt_DESeq[["DE_results_shrunken"]])

# Very few DE genes
M_globosa_Sc_Ch_DE_res <- M_globosa_Sc_Ch_filt_DESeq[["DE_results_shrunken_df"]] %>% dplyr::rename(feature=gene)


#Add annotations to DESeq2 results
M_globosa_Sc_Ch_DE_res_annot <- merge(M_globosa_Sc_Ch_DE_res,
                                        MGL_OG_annot %>% dplyr::select(gene,Description) %>% 
                                        dplyr::rename(feature=gene), 
                                        by = "feature")

```

# Cutibacterium acnes analysis

Get C_acnes counts after mapping to bacterial pangenomes.

```{r}

mtx_bt2_pangenome_counts <- lapply(mtx_to_pull, function(x){
  df <- read_tsv(file=paste0("../data/mtx_species_mapping/bt2_out/RNA/",x,"_bt2_microbe_pangenome_aligned_filtered_cov.tsv"),
                 show_col_types = FALSE)
  df$LIBID <- x
  
  return(df)
})


C_acnes_mtx_bt2_pangenome_counts <- lapply(mtx_bt2_pangenome_counts, function(x){
  
  output <-  calculate_species_TPM(x, species="Cutibacterium_acnes")
  
  return(output)
  
}) %>% do.call("rbind",.)

#add region information

C_acnes_mtx_bt2_pangenome_counts <- merge(C_acnes_mtx_bt2_pangenome_counts, mtx_stats_chosen %>% dplyr::select(LIBID,region), by ="LIBID")
                                          

#The worry is that these Scalp and cheeks are not comparable due to difference in C acnes coverage
#Sanity check shows this is ok

#Aggregate read counts per library

C_acnes_mtx_bt2_pangenome_counts_aggregate <- C_acnes_mtx_bt2_pangenome_counts %>% 
                                              group_by(LIBID,region) %>% summarise(total_read_count_in_sample=sum(unpaired_read_count_sum))

C_acnes_mtx_bt2_pangenome_counts_aggregate$region <- factor(C_acnes_mtx_bt2_pangenome_counts_aggregate$region,
                                                            levels=c("Sc","Ch","Vf","Ac","Tw"))

ggplot(C_acnes_mtx_bt2_pangenome_counts_aggregate %>% dplyr::filter(region %in% c("Sc","Ch")), 
       aes(x=region,y=log10(total_read_count_in_sample))) + geom_boxplot() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x=element_blank()) + 
  ggtitle("Expressed C. acnes features")


```
## Cutibacterium acnes comparisons between Scalp and cheek

```{r}

C_acnes_anno <- read_csv("../metadata/1747.Cutibacterium.acnes.pangenome.annotation.matrix.csv",
                         show_col_types = FALSE) %>% dplyr::rename(propan_clusterID=Clusters)
C_acnes_pangene_names <- read_tsv("../metadata/C_acnes_derep_pangenome_names.txt", 
                                  col_names = "pangene",
                                  show_col_types = FALSE) #9922


#Read counts for the propan clusters that are expressed by a majority of individuals with C acnes for Sc and Ch

C_acnes_mtx_cluster_counts <- C_acnes_mtx_bt2_pangenome_counts %>% 
  dplyr::select(LIBID, propan_clusterID, unpaired_read_count_sum)

#convert to input list
C_acnes_mtx_cluster_counts_list <- split(C_acnes_mtx_cluster_counts, 
                                                  f=C_acnes_mtx_cluster_counts$LIBID)

C_acnes_Sc_Ch_DE_QC <- QC_bacteria_RNA_sites(list_of_counts = C_acnes_mtx_cluster_counts_list,
                                                  species="Cutibacterium_acnes",
                                                  site="Sc", ref_site="Ch",
                                                  count_threshold = 200000,
                                                  shared_individuals_bet_sites=TRUE)

C_acnes_Sc_Ch_DE_QC[["PCA_plot"]]


# Minimum 200K C acnes reads
C_acnes_Sc_Ch_DE_analysis <- compare_bacteria_RNA_sites(list_of_counts = C_acnes_mtx_cluster_counts_list,
                                                        species_anno=C_acnes_anno, 
                                                        species_pangene_names = C_acnes_pangene_names,
                                                  species="Cutibacterium_acnes",
                                                  site="Sc", ref_site="Ch",
                                                  count_threshold = 200000,
                                                  shared_individuals_bet_sites=TRUE)

plotDispEsts(C_acnes_Sc_Ch_DE_analysis[["species_site_filt_DESeq"]]$dds)

DESeq2::plotMA(C_acnes_Sc_Ch_DE_analysis[["species_site_filt_DESeq"]]$DE_results_shrunken)


summary(C_acnes_Sc_Ch_DE_analysis[["species_site_filt_DESeq"]]$DE_results_shrunken)


#check the DE results 
C_acnes_Sc_Ch_DE_results_shrunken <- C_acnes_Sc_Ch_DE_analysis[["species_site_filt_DESeq"]]$DE_results_shrunken_df

C_acnes_Sc_Ch_DE_results_shrunken <- merge(C_acnes_Sc_Ch_DE_results_shrunken, C_acnes_anno, by = "propan_clusterID")

C_acnes_Sc_Ch_DE_results_shrunken_signif <- C_acnes_Sc_Ch_DE_results_shrunken %>% dplyr::filter(padj < 0.05)

```

## Panel C

Plot number of DE genes (upreg/downreg, adjusted p < 0.05) between scalp and cheek for M globosa, M restrica and C acnes


```{r}

microbial_Sc_vs_Ch_DE_list <- list(M_restricta_Sc_Ch_DE_res %>%  
                                     dplyr::rename(pangene=feature) %>% 
                                     dplyr::mutate(comparison="M_restricta"),
                                          M_globosa_Sc_Ch_DE_res %>% 
                                     dplyr::rename(pangene=feature) %>%
                                     dplyr::mutate(comparison="M_globosa"),
                                   C_acnes_Sc_Ch_DE_results_shrunken %>%
                                     dplyr::rename(pangene=propan_clusterID) %>%
                                            dplyr::mutate(comparison="C_acnes"))

microbial_Sc_vs_Ch_DE_tally <- lapply(microbial_Sc_vs_Ch_DE_list, function(x){
  
  output <- tally_signif_DE(x)
  return(output)
  
}) %>% do.call("rbind",.)


microbial_Sc_vs_Ch_DE_tally_fmt <- pivot_longer(data=microbial_Sc_vs_Ch_DE_tally,
                                   !comparison,
                                   names_to="direction",
                                   values_to="num_signif_features")

microbial_Sc_vs_Ch_DE_tally_fmt$direction <- gsub(microbial_Sc_vs_Ch_DE_tally_fmt$direction, pattern="num_",
                           replacement="")

microbial_Sc_vs_Ch_DE_tally_fmt$direction <- factor(microbial_Sc_vs_Ch_DE_tally_fmt$direction, 
                                       levels=c("upreg_features", "downreg_features"))

microbial_Sc_vs_Ch_DE_tally_barplot <- ggplot(microbial_Sc_vs_Ch_DE_tally_fmt, aes(x=comparison, y=num_signif_features, fill=direction)) +
  geom_col(position="dodge") + theme_classic()


ggsave(plot=microbial_Sc_vs_Ch_DE_tally_barplot, 
       filename = "../plots/microbial_Sc_vs_Ch_DE_tally_barplot.pdf")


microbial_Sc_vs_Ch_DE_tally_barplot


```


# Staphylococcus epidermidis analysis

## Data pre-processing for Staphylococcus epidermidis DESeq2 analysis

Make S epidermidis term2gene df for clusterprofiler (GSEA)

```{r}
#This is term2gene for KEGG pathways e.g. map03010
S_epi_KEGG_term2gene <- make_custom_term2gene_df(input_annot=S_epidermidis_anno,
                                                pangene_names = S_epidermidis_pangene_names)

#make term2gene for GO terms (MF and BP)

S_epi_GO_term2gene <- make_custom_term2gene_df(input_annot = eight_skin_bact_proteins_emapper_annot %>% 
                                           dplyr::filter(species=="Staphylococcus_epidermidis"), annotation_used = "GO")

```


Load kraken 2 counts for RNA 

```{r}

invivo_RNA_k2 <- lapply(mtx_to_pull, function(x){
  read_tsv(paste0("../data/kraken2/RNA/",x,"_merged_decont_k2.s.tsv"), 
           col_names = c("rel_abun","paired_reads","minimizer_count","distinct_minimizer_count","name"),
           show_col_types = FALSE) %>% dplyr::select(c("rel_abun", "paired_reads", "name"))
})

names(invivo_RNA_k2) <- mtx_to_pull

invivo_RNA_k2 <- lapply(mtx_to_pull, function(x){
  invivo_RNA_k2[[x]]$LIBID <- x
  return(invivo_RNA_k2[[x]])
})

invivo_RNA_k2_df <- do.call("rbind", invivo_RNA_k2)

names(invivo_RNA_k2) <- mtx_to_pull

```

Select in vivo samples with a minimum threshold (200000 reads) of microbial RNA (e.g. S. epidermidis or C. acnes depending on the analysis). 

```{r}

S_epi_in_vivo_meta <- species_threshold_fn(species="Staphylococcus epidermidis", 
                                           paired_read_threshold = 200000,
                                           k2_input=invivo_RNA_k2_df)

S_epi_in_vivo_meta <- merge(S_epi_in_vivo_meta %>% dplyr::rename(mtx_LIBID=LIBID), mtx_mgx_stats_chosen, by = "mtx_LIBID")
#S_epi_in_vivo_meta <- S_epi_in_vivo_meta %>% dplyr::filter(region=="Ch"|region=="Tw")

#to increase number of samples, group sebaceous sites together (Ch and Sc) for S epidermidis
S_epi_in_vivo_meta$region_fmt <- ifelse(S_epi_in_vivo_meta$region == "Tw", "Tw", "Sebaceous" )



#For C acnes
C_acnes_in_vivo_meta <- species_threshold_fn(species="Cutibacterium acnes", paired_read_threshold = 200000, k2_input=invivo_RNA_k2_df)
C_acnes_in_vivo_meta <- merge(C_acnes_in_vivo_meta %>% dplyr::rename(mtx_LIBID=LIBID), mtx_mgx_stats_chosen, by = "mtx_LIBID")


```

prepare DESeq2 metadata for S epidermidis

```{r}

temp_meta <- S_epi_in_vitro_metadata %>% dplyr::select(Library_ID, expt_desc, study) %>% 
  unique %>% 
  dplyr::rename(condition=expt_desc) %>% as.data.frame()
temp_meta$expt_class <- "in_vitro"

temp_meta2 <- S_epi_in_vivo_meta %>% dplyr::select(mtx_LIBID, region_fmt) %>% dplyr::rename(condition=region_fmt)
temp_meta2$expt_class <- "in_vivo"
temp_meta2$study <- "this_study"

in_vitro_LIBID <- temp_meta %>% pull(Library_ID) 

#Create row names with the associated sample names

rownames(temp_meta) <- temp_meta$Library_ID
rownames(temp_meta2) <- temp_meta2$mtx_LIBID

DESeq_metadata <- rbind(temp_meta %>% dplyr::select(-Library_ID), temp_meta2 %>% dplyr::select(-mtx_LIBID))

#batch information within the "study" variable
DESeq_metadata$study <- factor(DESeq_metadata$study)

#DESeq_metadata$batch <- ifelse(str_detect(rownames(DESeq_metadata),pattern="Sepi|MHS"),
                               #"this_study","other_studies")

#DESeq_metadata$batch <- factor(DESeq_metadata$batch)

```


Reminder: Both in vivo and in vitro RNA seq reads were mapped to the eight skin microbial pangenome using bowtie2.

Prepare dataset of raw counts per cluster for S epidermidis. This will be for DESeq analysis and for deriving inputs for metabolic modelling.

```{r}

invivo_mtx_counts <- lapply(S_epi_in_vivo_meta$mtx_LIBID, function(x){
  df <- read_tsv(file=paste0("../data/mtx_species_mapping/bt2_out/RNA/",x,"_bt2_microbe_pangenome_aligned_filtered_cov.tsv"),
                 show_col_types = FALSE)
  df$LIBID <- x
  
  df$condition <- DESeq_metadata[x, "condition"]
  
  #select only S epidermidis gene features
  
  df_filt <- df %>% dplyr::filter(str_detect(df$pangene, pattern = "Staphylococcus_epidermidis"))
  
  #sum up read counts within the propan clusters
  ##clusters fit the propan definitions
  df_filt$propan_clusterID <- str_match(string=df_filt$pangene, pattern="cluster_[0-9]*")[,1]
  
  df_filt_clust <- df_filt %>% group_by(propan_clusterID) %>% 
    mutate(unpaired_read_count_sum=sum(unpaired_read_count)) %>% ungroup()
  
  output <- df_filt_clust %>% dplyr::select(c("propan_clusterID","unpaired_read_count_sum", "LIBID")) %>% unique(.)
  
  output <- output %>% dplyr::rename(pangene=propan_clusterID,unpaired_read_count=unpaired_read_count_sum)
  
  return(output)
})


invitro_mtx_counts <- lapply(in_vitro_LIBID, function(x){
  df <- read_tsv(file=paste0("../data/mtx_species_mapping/S_epi_RNAseq_align_out/bt2_out/",x,"_bt2_microbe_pangenome_aligned_filtered_cov.tsv"),
                 show_col_types = FALSE)
  df$LIBID <- x
  
  df$condition <- DESeq_metadata[x, "condition"]
  
  #select only S epidermidis gene features
  
  df_filt <- df %>% dplyr::filter(str_detect(df$pangene, pattern = "Staphylococcus_epidermidis"))
  
  #sum up read counts within the propan clusters
  ##clusters fit the propan definitions
  df_filt$propan_clusterID <- str_match(string=df_filt$pangene, pattern="cluster_[0-9]*")[,1]
  
  df_filt_clust <- df_filt %>% group_by(propan_clusterID) %>% 
    mutate(unpaired_read_count_sum=sum(unpaired_read_count)) %>% ungroup()
  
  output <- df_filt_clust %>% dplyr::select(c("propan_clusterID","unpaired_read_count_sum", "LIBID")) %>% unique(.)
  
  output <- output %>% dplyr::rename(pangene=propan_clusterID,unpaired_read_count=unpaired_read_count_sum)
  
  return(output)
})

all_S_epi_mtx_counts <- c(invivo_mtx_counts, invitro_mtx_counts)

#https://stackoverflow.com/questions/72040033/name-lists-elements-based-on-column-they-are-containing
names(all_S_epi_mtx_counts) <- sapply(all_S_epi_mtx_counts, function(x) unique(x$LIBID))

```


# Run DESeq2 for in vivo versus in vitro comparisons (Staphylococcus epidermidis)

## Comparison 1: Tw vs stationary phase

```{r}

#compare with in house data and three external studies: Avican et al 2021, Ahle et al 2022 and Wang et al 2017

Tw_vs_all_stat_phase <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                                                 lib_vec = get_lib_vec_fn(in_vitro_condition="stationary_phase_liquid", 
                                                                          in_vitro_study = c("this_study", 
                                                                                             "Avican_et_al_2021", 
                                                                                             "Wang_et_al_2017")), 
                                              run_test = TRUE, batch_correct = TRUE)

#PCA should be visualized after limma::removebatcheffect instead. See previous code chunk,
#plotPCA(Tw_vs_all_stat_phase[["vsd"]], intgroup = "condition", returnData=FALSE)


plotDispEsts(Tw_vs_all_stat_phase[["dds"]])

DESeq2::plotMA(Tw_vs_all_stat_phase[["DE_results_shrunken"]])

boxplot(Tw_vs_all_stat_phase$vsd_mat, 
         xlab="", 
         ylab="VST counts")


summary(Tw_vs_all_stat_phase[["DE_results_shrunken"]])

#save results
Tw_vs_all_stat_phase_DE_res <- Tw_vs_all_stat_phase[["DE_results_shrunken_df"]]

Tw_vs_all_stat_phase_DE_res <- merge(Tw_vs_all_stat_phase_DE_res,
                             S_epidermidis_anno %>% dplyr::rename(pangene=propan_clusterID))


write_tsv(Tw_vs_all_stat_phase_DE_res,"../data/DESeq2_out/S_epidermidis/S_epi_Tw_vs_all_stat_phase_DE_res.tsv")


##############################
#GENE SET ENRICHMENT ANALYSIS#
##############################

#GSEA MF
#No GO terms enriched for MF under specific p value cut off
S_epi_Tw_vs_all_stat_phase_DE_GSEA_GO_MF <- GSEA_from_DESeq(genelist = Tw_vs_all_stat_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_stat_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_MF_term2name$from ),
                                         term2name = GO_MF_term2name)


#GSEA BP
S_epi_Tw_vs_all_stat_phase_DE_GSEA_GO_BP <- GSEA_from_DESeq(genelist = Tw_vs_all_stat_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_stat_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_BP_term2name$from ),
                                         term2name = GO_BP_term2name)

#How about KEGG pathways
Tw_vs_all_stat_phase_GSEA_KEGG <- GSEA_from_DESeq(genelist = Tw_vs_all_stat_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_stat_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_KEGG_term2gene,
                                         term2name = KEGG_term2name)

clusterProfiler::dotplot(Tw_vs_all_stat_phase_GSEA_KEGG, showCategory=10, split=".sign") + facet_grid(.~.sign)



S_epi_Tw_vs_all_stat_phase_DE_GSEA_GO_BP_df <- S_epi_Tw_vs_all_stat_phase_DE_GSEA_GO_BP@result
S_epi_Tw_vs_all_stat_phase_DE_GSEA_GO_BP_df$set_annotation <- "GO_BP" 

S_epi_Tw_vs_all_stat_phase_DE_GSEA_KEGG_df <- Tw_vs_all_stat_phase_GSEA_KEGG@result
S_epi_Tw_vs_all_stat_phase_DE_GSEA_KEGG_df$set_annotation <- "KEGG"

write_tsv(rbind(S_epi_Tw_vs_all_stat_phase_DE_GSEA_GO_BP_df,
                S_epi_Tw_vs_all_stat_phase_DE_GSEA_KEGG_df),
          "../data/DESeq2_out/S_epi_Tw_vs_all_stat_phase_GSEA.tsv")

S_epi_Tw_vs_all_stat_phase_GSEA <- read_tsv("../data/DESeq2_out/S_epi_Tw_vs_all_stat_phase_GSEA.tsv", show_col_types = FALSE)

```

##Comparison 2: Tw vs osmotic stress in vitro (in house and external studies). 


```{r}

#OS: osmotic stress

Tw_vs_all_OS <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                                                 lib_vec = get_lib_vec_fn(in_vitro_condition="osmotic_stress_liquid", in_vitro_study = c("this_study", "Avican_et_al_2021", "Wang_et_al_2017")), 
                                                 batch_correct=TRUE, 
                                                 run_test = TRUE)


plotDispEsts(Tw_vs_all_OS[["dds"]])

DESeq2::plotMA(Tw_vs_all_OS[["DE_results_shrunken"]])


summary(Tw_vs_all_OS[["DE_results_shrunken"]])

Tw_vs_all_OS_DE_res <- Tw_vs_all_OS$DE_results_shrunken_df
Tw_vs_all_OS_DE_res <- merge(Tw_vs_all_OS_DE_res,
                             S_epidermidis_anno %>% dplyr::rename(pangene=propan_clusterID))

write_tsv(Tw_vs_all_OS_DE_res,"../data/DESeq2_out/S_epidermidis/S_epi_Tw_vs_all_OS_DE_res.tsv")

##############################
#GENE SET ENRICHMENT ANALYSIS#
##############################

#GSEA MF
S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF <- GSEA_from_DESeq(genelist = Tw_vs_all_OS$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_OS$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_MF_term2name$from ),
                                         term2name = GO_MF_term2name)


clusterProfiler::dotplot(S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF, showCategory=10, split=".sign") + facet_grid(.~.sign)

#associate with parent GO slim (metagenomes)
S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF_slim <- associate_GO_to_slim(S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF@result$ID, mode="MF")

S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF_fmt <- prepare_NES_plot_input(df=S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF@result,
                                                               slimdf=S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF_slim,
                                                               chosen_parent_ids = "transporter activity")


S_epi_Tw_vs_all_OS_GSEA_slim_NES_barplot <- make_NES_barplot_with_slim(S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF_fmt, title= "S. epidermidis transcripts, TW vs Osmotic Stress, MF")

ggsave(filename="../plots/S_epi_Tw_vs_all_OS_GSEA_slim_NES_barplot.pdf",
       plot=S_epi_Tw_vs_all_OS_GSEA_slim_NES_barplot)

S_epi_Tw_vs_all_OS_GSEA_slim_NES_barplot


S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF_df <- S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF@result
S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF_df$set_annotation <- "GO_MF" 

write_tsv(S_epi_Tw_vs_all_OS_DE_GSEA_GO_MF_df,
          "../data/DESeq2_out/S_epi_Tw_vs_all_OS_GSEA.tsv")


S_epi_Tw_vs_all_OS_GSEA <- read_tsv("../data/DESeq2_out/S_epi_Tw_vs_all_OS_GSEA.tsv", show_col_types = FALSE)



#BP, no hits
S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP <- GSEA_from_DESeq(genelist = Tw_vs_all_OS$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_OS$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_BP_term2name$from ),
                                         term2name = GO_BP_term2name)


#clusterProfiler::dotplot(S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP, showCategory=10, split=".sign") + facet_grid(.~.sign)

#associate with parent GO slim (metagenomes)
#S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP_slim <- associate_GO_to_slim(S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP@result$ID, mode="BP")

#S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP_fmt <- prepare_NES_plot_input(df=S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP@result,
#                                                               slimdf=S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP_slim,
#                                                               chosen_parent_ids = "transport")

#make_NES_barplot_with_slim(S_epi_Tw_vs_all_OS_DE_GSEA_GO_BP_fmt, title= "S. epidermidis transcripts, TW vs Osmotic Stress, BP")


#No KEGG GSEA enrichments
Tw_vs_all_OS_GSEA_KEGG <- GSEA_from_DESeq(genelist = Tw_vs_all_OS$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_OS$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_KEGG_term2gene,
                                         term2name = KEGG_term2name)


```
##Comparison 3: Tw vs log phase in vitro

Do GSEA of Tw vs all (in house + external) log phase in vitro


```{r}

Tw_vs_all_log_phase <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                                                 lib_vec = get_lib_vec_fn(in_vitro_condition="log_phase_liquid", 
                                                                          in_vitro_study = c("this_study", 
                                                                                             "Avican_et_al_2021", 
                                                                                             "Wang_et_al_2017")), 
                                             run_test = TRUE, batch_correct = TRUE)


#plotPCA(Tw_vs_all_log_phase[["vsd"]], intgroup = "condition", returnData=FALSE)

plotDispEsts(Tw_vs_all_log_phase[["dds"]])

DESeq2::plotMA(Tw_vs_all_log_phase[["DE_results_shrunken"]])

summary(Tw_vs_all_log_phase[["DE_results_shrunken"]])

Tw_vs_all_log_phase_DE_res <- Tw_vs_all_log_phase$DE_results_shrunken_df
Tw_vs_all_log_phase_DE_res <- merge(Tw_vs_all_log_phase_DE_res,
                             S_epidermidis_anno %>% dplyr::rename(pangene=propan_clusterID))

write_tsv(Tw_vs_all_log_phase_DE_res,"../data/DESeq2_out/S_epidermidis/S_epi_Tw_vs_all_log_phase_DE_res.tsv")


##############################
#GENE SET ENRICHMENT ANALYSIS#
##############################

#GSEA MF

#MF no hits
S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF <- GSEA_from_DESeq(genelist = Tw_vs_all_log_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_log_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_MF_term2name$from ),
                                         term2name = GO_MF_term2name)


#clusterProfiler::dotplot(S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF, showCategory=10, split=".sign") + facet_grid(.~.sign)

#associate with parent GO slim (metagenomes)
#S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF_slim <- associate_GO_to_slim(S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF@result$ID, mode="MF")

#S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF_fmt <- prepare_NES_plot_input(df=S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF@result,
#                       slimdf=S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF_slim,
#                       chosen_parent_ids = "transporter activity")


#Note: the annotation of GO:0042626 was missed out from the *obo file. Manually fix the pdf file to reflect the correct category
#make_NES_barplot_with_slim(S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF_fmt, title= "S. epidermidis transcripts, TW vs log phase, MF")

#ggsave(filename="../plots/DESeq_S_epi/S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_MF.pdf",
#       plot=last_plot(),
#       height = 8, width = 10)


#GSEA BP

S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP <- GSEA_from_DESeq(genelist = Tw_vs_all_log_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_log_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_BP_term2name$from ),
                                         term2name = GO_BP_term2name)


clusterProfiler::dotplot(S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP, showCategory=10, split=".sign") + facet_grid(.~.sign)

S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP_df <- S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP@result
S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP_df$set_annotation <- "GO_BP" 

write_tsv(S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP_df,
          "../data/DESeq2_out/S_epi_Tw_vs_all_log_phase_GSEA.tsv")


S_epi_Tw_vs_all_log_phase_GSEA <- read_tsv("../data/DESeq2_out/S_epi_Tw_vs_all_log_phase_GSEA.tsv", show_col_types = FALSE)


#No KEGG GSEA enrichments

Tw_vs_all_log_phase_GSEA_KEGG <- GSEA_from_DESeq(genelist = Tw_vs_all_log_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Tw_vs_all_log_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_KEGG_term2gene,
                                         term2name = KEGG_term2name)


#Associate GO terms with slimmed parental GO (metagenomes)
S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP_slim <- associate_GO_to_slim(S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP@result$ID, mode="BP")


S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP_fmt <- prepare_NES_plot_input(df=S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP@result,
                       slimdf=S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP_slim,
                       chosen_parent_ids = c("carbohydrate metabolic process", 
                       "small molecule metabolic process", 
                       "transport"))

S_epi_Tw_vs_all_log_phase_GSEA_slim_NES_barplot <- make_NES_barplot_with_slim(S_epi_Tw_vs_all_log_phase_DE_GSEA_GO_BP_fmt, title= "S. epidermidis transcripts, TW vs log phase")

ggsave(filename="../plots/S_epi_Tw_vs_all_log_phase_GSEA_slim_NES_barplot.pdf",
       plot=S_epi_Tw_vs_all_log_phase_GSEA_slim_NES_barplot,
       height = 8, width = 10)


S_epi_Tw_vs_all_log_phase_GSEA_slim_NES_barplot

```
## Comparison 4: Seb vs OS in vitro

Note cluster_17808

```{r}
#all studies

Seb_vs_all_OS <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                      lib_vec = get_lib_vec_fn(in_vivo_site = "Sebaceous",                                                        in_vitro_condition="osmotic_stress_liquid", 
                                    in_vitro_study = c("this_study", 
                                   "Avican_et_al_2021", 
                                   "Wang_et_al_2017")), 
                      run_test = TRUE,
                      batch_correct = TRUE)



#plotPCA(Seb_vs_all_OS[["vsd"]], intgroup = "condition", returnData=FALSE)


Seb_vs_all_OS_DE_res <- Seb_vs_all_OS$DE_results_shrunken_df
Seb_vs_all_OS_DE_res <- merge(Seb_vs_all_OS_DE_res,
                             S_epidermidis_anno %>% dplyr::rename(pangene=propan_clusterID))


write_tsv(Seb_vs_all_OS_DE_res,"../data/DESeq2_out/S_epi_Seb_vs_all_OS_DE_res.tsv")


##############################
#GENE SET ENRICHMENT ANALYSIS#
##############################

#GSEA MF
S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF <- GSEA_from_DESeq(genelist = Seb_vs_all_OS$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_OS$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_MF_term2name$from ),
                                         term2name = GO_MF_term2name)


clusterProfiler::dotplot(S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF, showCategory=10, split=".sign") + facet_grid(.~.sign)

#Associate GO terms with slimmed parental GO (metagenomes)


S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF_slim <- associate_GO_to_slim(S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF@result$ID, mode="MF")


S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF_fmt <- prepare_NES_plot_input(df=S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF@result,
                       slimdf=S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF_slim,
                       chosen_parent_ids = c("transporter activity",
											"hydrolase activity", 
											"ion binding", 
											"nucleotide binding",
											"nucleic acid binding",
											"oxidoreductase activity",
											"transferase activity",
											"structural constituent of ribosome"))


make_NES_barplot_with_slim(S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF_fmt, title= "S. epidermidis transcripts, Sebaceous vs osmotic stress, MF")


#GSEA BP

S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP <- GSEA_from_DESeq(genelist = Seb_vs_all_OS$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_OS$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_BP_term2name$from ),
                                         term2name = GO_BP_term2name)


clusterProfiler::dotplot(S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP, showCategory=10, split=".sign") + facet_grid(.~.sign)

#Associate GO terms with slimmed parental GO (metagenomes)

S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP_slim <- associate_GO_to_slim(S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP@result$ID, mode="BP")

S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP_fmt <- prepare_NES_plot_input(df=S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP@result,
                       slimdf=S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP_slim,
                       chosen_parent_ids = c("small molecule metabolic process",
											"transport", 
											"carbohydrate metabolic process", 
											"translation", 
											"phosphorylation"))


make_NES_barplot_with_slim(S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP_fmt, title= "S. epidermidis transcripts, Sebaceous vs osmotic stress, BP")



#GSEA KEGG

Seb_vs_all_OS_GSEA_KEGG <- GSEA_from_DESeq(genelist = Seb_vs_all_OS$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_OS$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_KEGG_term2gene,
                                         term2name = KEGG_term2name)

clusterProfiler::dotplot(Seb_vs_all_OS_GSEA_KEGG, showCategory=10, split=".sign") + facet_grid(.~.sign)


S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF_df <- S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF@result
S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF_df$set_annotation <- "GO_MF" 


S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP_df <- S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP@result
S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP_df$set_annotation <- "GO_BP" 

S_epi_Seb_vs_all_OS_DE_GSEA_KEGG_df <- Seb_vs_all_OS_GSEA_KEGG@result
S_epi_Seb_vs_all_OS_DE_GSEA_KEGG_df$set_annotation <- "KEGG"

write_tsv(rbind(S_epi_Seb_vs_all_OS_DE_GSEA_GO_MF_df,
                S_epi_Seb_vs_all_OS_DE_GSEA_GO_BP_df,
                S_epi_Seb_vs_all_OS_DE_GSEA_KEGG_df), "../data/DESeq2_out/S_epi_Seb_vs_all_OS_GSEA.tsv")


S_epi_Seb_vs_all_OS_GSEA <- read_tsv("../data/DESeq2_out/S_epi_Seb_vs_all_OS_GSEA.tsv", show_col_types = FALSE)




```

## Comparison 5: Seb vs log phase in vitro

Sebaceous vs external and in-house (all) log phase

```{r}

Seb_vs_all_log_phase <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                      lib_vec = get_lib_vec_fn(in_vivo_site = "Sebaceous",                                                        in_vitro_condition="log_phase_liquid", 
                                    in_vitro_study = c("this_study", 
                                   "Avican_et_al_2021", 
                                   "Wang_et_al_2017")), 
                      run_test = TRUE,
                      batch_correct = TRUE)



#plotPCA(Seb_vs_all_log_phase[["vsd"]], intgroup = "condition", returnData=FALSE)

Seb_vs_all_log_phase_DE_res <- Seb_vs_all_log_phase$DE_results_shrunken_df
Seb_vs_all_log_phase_DE_res <- merge(Seb_vs_all_log_phase_DE_res,
                             S_epidermidis_anno %>% dplyr::rename(pangene=propan_clusterID))

write_tsv(Seb_vs_all_log_phase_DE_res,"../data/DESeq2_out/S_epi_Seb_vs_all_log_phase_DE_res.tsv")

##############################
#GENE SET ENRICHMENT ANALYSIS#
##############################

#GSEA MF
S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF <- GSEA_from_DESeq(genelist = Seb_vs_all_log_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_log_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_MF_term2name$from ),
                                         term2name = GO_MF_term2name)


clusterProfiler::dotplot(S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF, showCategory=10, split=".sign") + facet_grid(.~.sign)


#Associate GO terms with slimmed parental GO (metagenomes)

S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF_slim <- associate_GO_to_slim(S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF@result$ID, mode="MF")

S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF_fmt <- prepare_NES_plot_input(df=S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF@result,
                       slimdf=S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF_slim,
                       chosen_parent_ids = c("catalytic activity",
                      "nucleic acid binding",
											"transporter activity", 
											"transferase activity", 
											"structural constituent of ribosome", 
											"metal ion binding"))


make_NES_barplot_with_slim(S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF_fmt, title= "S. epidermidis transcripts, Sebaceous vs log phase, MF")


#GSEA BP
S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP <- GSEA_from_DESeq(genelist = Seb_vs_all_log_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_log_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_BP_term2name$from ),
                                         term2name = GO_BP_term2name)


clusterProfiler::dotplot(S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP, showCategory=10, split=".sign") + facet_grid(.~.sign)

#Associate GO terms with slimmed parental GO (metagenomes)

S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP_slim <- associate_GO_to_slim(S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP@result$ID, mode="BP")

S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP_fmt <- prepare_NES_plot_input(df=S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP@result,
                                                                       slimdf=S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP_slim,
  chosen_parent_ids = c("nitrogen compound metabolic process",
											"small molecule metabolic process", 
											"carbohydrate metabolic process", 
											"RNA metabolic process", 
											"transport",
											"phosphorylation"))

make_NES_barplot_with_slim(S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP_fmt, title= "S. epidermidis transcripts, Sebaceous vs log phase, BP")



#GSEA KEGG
Seb_vs_all_log_phase_GSEA_KEGG <- GSEA_from_DESeq(genelist = Seb_vs_all_log_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_log_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_KEGG_term2gene,
                                         term2name = KEGG_term2name)

clusterProfiler::dotplot(Seb_vs_all_log_phase_GSEA_KEGG, showCategory=10, split=".sign") + facet_grid(.~.sign)


S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF_df <- S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF@result
S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF_df$set_annotation <- "GO_MF" 


S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP_df <- S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP@result
S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP_df$set_annotation <- "GO_BP" 

S_epi_Seb_vs_all_log_phase_DE_GSEA_KEGG_df <- Seb_vs_all_log_phase_GSEA_KEGG@result
S_epi_Seb_vs_all_log_phase_DE_GSEA_KEGG_df$set_annotation <- "KEGG"

write_tsv(rbind(S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_MF_df,
  S_epi_Seb_vs_all_log_phase_DE_GSEA_GO_BP_df,
                S_epi_Seb_vs_all_log_phase_DE_GSEA_KEGG_df), "../data/DESeq2_out/S_epi_Seb_vs_all_log_phase_GSEA.tsv")


S_epi_Seb_vs_all_log_phase_GSEA <- read_tsv("../data/DESeq2_out/S_epi_Seb_vs_all_log_phase_GSEA.tsv", show_col_types = FALSE)

```

## Comparison 6: Seb vs stat phase in vitro

```{r}

Seb_vs_all_stat_phase <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                                                 lib_vec = get_lib_vec_fn(in_vivo_site = "Sebaceous", 
                                                                          in_vitro_condition="stationary_phase_liquid", 
                                                                          in_vitro_study = c("this_study", 
                                                                                             "Avican_et_al_2021", 
                                                                                             "Wang_et_al_2017")), 
                                               run_test = TRUE, batch_correct = TRUE)


#plotPCA(Seb_vs_all_stat_phase[["vsd"]], intgroup = "condition", returnData=FALSE)


plotDispEsts(Seb_vs_all_stat_phase[["dds"]])

DESeq2::plotMA(Seb_vs_all_stat_phase[["DE_results_shrunken"]])


summary(Seb_vs_all_stat_phase[["DE_results_shrunken"]])



Seb_vs_all_stat_phase_DE_res <- Seb_vs_all_stat_phase[["DE_results_shrunken_df"]]

Seb_vs_all_stat_phase_DE_res <- merge(Seb_vs_all_stat_phase_DE_res,
                             S_epidermidis_anno %>% dplyr::rename(pangene=propan_clusterID))


write_tsv(Seb_vs_all_stat_phase_DE_res,"../data/DESeq2_out/S_epi_Seb_vs_all_stat_phase_DE_res.tsv")

##############################
#GENE SET ENRICHMENT ANALYSIS#
##############################

#GSEA MF
S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_MF <- GSEA_from_DESeq(genelist = Seb_vs_all_stat_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_stat_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_MF_term2name$from ),
                                         term2name = GO_MF_term2name)


clusterProfiler::dotplot(S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_MF, showCategory=10, split=".sign") + facet_grid(.~.sign)


#GSEA BP
S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_BP <- GSEA_from_DESeq(genelist = Seb_vs_all_stat_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_stat_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_GO_term2gene %>% dplyr::filter(from %in% GO_BP_term2name$from ),
                                         term2name = GO_BP_term2name)


clusterProfiler::dotplot(S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_BP, showCategory=10, split=".sign") + facet_grid(.~.sign)


#GSEA KEGG
Seb_vs_all_stat_phase_GSEA_KEGG <- GSEA_from_DESeq(genelist = Seb_vs_all_stat_phase$DE_results_shrunken_df$log2FoldChange,
                                         gene_names = Seb_vs_all_stat_phase$DE_results_shrunken_df$pangene,
                                         term2gene = S_epi_KEGG_term2gene,
                                         term2name = KEGG_term2name)

clusterProfiler::dotplot(Seb_vs_all_stat_phase_GSEA_KEGG, showCategory=10, split=".sign") + facet_grid(.~.sign)



S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_MF_df <- S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_MF@result
S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_MF_df$set_annotation <- "GO_MF" 


S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_BP_df <- S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_BP@result
S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_BP_df$set_annotation <- "GO_BP" 

S_epi_Seb_vs_all_stat_phase_DE_GSEA_KEGG_df <- Seb_vs_all_stat_phase_GSEA_KEGG@result
S_epi_Seb_vs_all_stat_phase_DE_GSEA_KEGG_df$set_annotation <- "KEGG"

write_tsv(rbind(S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_MF_df,
  S_epi_Seb_vs_all_stat_phase_DE_GSEA_GO_BP_df,
                S_epi_Seb_vs_all_stat_phase_DE_GSEA_KEGG_df),
          "../data/DESeq2_out/S_epi_Seb_vs_all_stat_phase_GSEA.tsv")


S_epi_Seb_vs_all_stat_phase_GSEA <- read_tsv("../data/DESeq2_out/S_epi_Seb_vs_all_stat_phase_GSEA.tsv", show_col_types = FALSE)

```


## Panel E (Left)

Principal component analysis (PCA) plot of S. epidermidis gene expression signatures from various in vitro and in vivo samples after batch correction using Limma. 

Caveat: This analysis is solely for creating PCA plots. 
https://support.bioconductor.org/p/133203/

```{r}

#Remove "Ahle et al 2022" from the metadata because strains were grown on plates and not on liquid.

DESeq_metadata <- DESeq_metadata %>% dplyr::filter(study != "Ahle_et_al_2022")
DESeq_metadata$study <- factor(DESeq_metadata$study, levels=c("Wang_et_al_2017","Avican_et_al_2021", "this_study"))



#batch_correct = FALSE here because the design matrix shouldn't include the batch effect for purposes of PCA.
S_epi_in_vivo_vs_in_vitro_DE <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                                                 lib_vec = get_lib_vec_fn(in_vivo_site = c("Sebaceous",
                                                                                           "Tw"),
                                                                          in_vitro_condition=c("stationary_phase_liquid","osmotic_stress_liquid", "log_phase_liquid"), in_vitro_study = c("this_study", "Avican_et_al_2021", "Wang_et_al_2017")), 
                                                 batch_correct = FALSE, run_test = FALSE)


assay(S_epi_in_vivo_vs_in_vitro_DE[["vsd"]])<- limma::removeBatchEffect(assay(S_epi_in_vivo_vs_in_vitro_DE[["vsd"]]), 
                                                                         batch=S_epi_in_vivo_vs_in_vitro_DE[["selected_metadata"]]$study,
                                                                         design= model.matrix(~condition, 
                                                                                              data = S_epi_in_vivo_vs_in_vitro_DE[["selected_metadata"]]))


#PCA plot after batch correction
plotPCA(S_epi_in_vivo_vs_in_vitro_DE[["vsd"]], intgroup = "expt_class", returnData=FALSE)


invivo_invitro_PCA_batch_corrected <- plotPCA(S_epi_in_vivo_vs_in_vitro_DE[["vsd"]], intgroup = "condition", returnData=FALSE)


invivo_invitro_PCA_batch_corrected <- invivo_invitro_PCA_batch_corrected + scale_color_manual(values=c("dodgerblue2",
                                                                 "#E31A1C",
                                                                 "green4","black","#6A3D9A"))

ggsave(filename="../plots/invivo_invitro_PCA_batch_corrected.pdf", plot=invivo_invitro_PCA_batch_corrected)

invivo_invitro_PCA_batch_corrected

```


```{r}
##How about keeping the same principal components and just visualizing the in vitro data. How do they separate?



invivo_invitro_PCA_batch_corrected_data <- plotPCA(S_epi_in_vivo_vs_in_vitro_DE[["vsd"]], intgroup = "condition", returnData=TRUE)

#add study metadata again
invivo_invitro_PCA_batch_corrected_df <- merge(invivo_invitro_PCA_batch_corrected_data[[1]],
                                                   S_epi_in_vitro_metadata %>% dplyr::select(Library_ID,study) %>% dplyr::rename("name"="Library_ID"),
                                                   by = "name", all.x=TRUE)


invivo_invitro_PCA_batch_corrected_df$study <- ifelse(is.na(invivo_invitro_PCA_batch_corrected_df$study), "this_study",
                                                      invivo_invitro_PCA_batch_corrected_df$study)

invivo_invitro_PCA_batch_corrected_zoomed <- ggplot(data=invivo_invitro_PCA_batch_corrected_df %>% dplyr::filter(!group %in% c("Sebaceous",
                                                                                 "Tw")), 
       aes_string(x="PC1", y="PC2", color="group")) + geom_point(aes(shape=study),size=5) + 
    xlab("PC1") +
      ylab("PC2") +
        coord_fixed() + theme_classic() + scale_color_manual(values=c("dodgerblue2",  "#E31A1C","black")) +
  scale_shape_manual(values=c(15, 17, 4)) + scale_y_break(c(-20,-40))



invivo_invitro_PCA_batch_corrected_zoomed

ggsave(filename="../plots/invivo_invitro_PCA_batch_corrected_zoomed.pdf", plot=invivo_invitro_PCA_batch_corrected_zoomed)

invivo_invitro_PCA_batch_corrected_zoomed

```

##Panel E (right)

Barplots showing the number of differentially expressed genes from in vivo S. epidermidis versus different in vitro conditions. Sebaceous sites refer to both scalp and cheek samples with  200,000 S. epidermidis reads per library (n = 6). Toe webs refer to toe web samples with  200,000 S. epidermidis reads per library (n = 12)

Bar chart of number of upreg and downreg genes for each comparison (6 in total)

p adj < 0.05

```{r}

Seb_vs_all_log_phase_DE_res$comparison <- "Seb_vs_log_phase"
Seb_vs_all_OS_DE_res$comparison <- "Seb_vs_OS"
Seb_vs_all_stat_phase_DE_res$comparison <- "Seb_vs_stat_phase"
Tw_vs_all_log_phase_DE_res$comparison <- "Tw_vs_log_phase"
Tw_vs_all_OS_DE_res$comparison <- "Tw_vs_OS"
Tw_vs_all_stat_phase_DE_res$comparison <- "Tw_vs_stat_phase"


S_epi_DE_list <- tibble::lst(Seb_vs_all_log_phase_DE_res,
                             Seb_vs_all_OS_DE_res,
                             Seb_vs_all_stat_phase_DE_res,
                             Tw_vs_all_log_phase_DE_res,
                             Tw_vs_all_OS_DE_res,
                             Tw_vs_all_stat_phase_DE_res)


S_epi_DE_tally <- lapply(S_epi_DE_list, function(x){
  
  output <- tally_signif_DE(x)
  return(output)
  
}) %>% do.call("rbind",.)


S_epi_DE_tally_fmt <- pivot_longer(data=S_epi_DE_tally,
                                   !comparison,
                                   names_to="direction",
                                   values_to="num_signif_features")

S_epi_DE_tally_fmt$direction <- gsub(S_epi_DE_tally_fmt$direction, pattern="num_",
                           replacement="")

S_epi_DE_tally_fmt$direction <- factor(S_epi_DE_tally_fmt$direction, 
                                       levels=c("upreg_features", "downreg_features"))

S_epi_DE_tally_barplot <- ggplot(S_epi_DE_tally_fmt, aes(x=comparison, y=num_signif_features, fill=direction)) +
  geom_col(position="dodge") + theme_classic()


ggsave(plot=S_epi_DE_tally_barplot, 
       filename = "../plots/S_epi_DE_tally_barplot.pdf")


S_epi_DE_tally_barplot 
```

# Supplementary figure NES plots


Enriched gene sets in Staphylococcus epidermidis transcriptomes for sebaceous sites versus in vitro conditions. 

Barplots of normalized enrichment scores derived from gene set enrichment analysis of differentially expressed genes between sebaceous (scalp and cheek) sites (n=6) and various in vitro conditions (n=9, 6, 9 for log phase, osmotic stress & stationary phase respectively). 

Only gene sets or pathways which were significantly enriched in all 3 comparisons are displayed here. Libraries representing sebaceous sites had 200,000 S. epidermidis reads each.

Y axis are pathways/gene sets. X axis are normalized enrichment score, faceted by GO_MF, GO_BP and KEGG

```{r}


S_epi_Seb_vs_all_log_phase_GSEA$comparison <- "Seb_vs_log_phase"

S_epi_Seb_vs_all_OS_GSEA$comparison <- "Seb_vs_osmotic_stress"

S_epi_Seb_vs_all_stat_phase_GSEA$comparison <- "Seb_vs_stationary_phase"



Seb_vs_in_vitro_shared_gene_sets <- Reduce(intersect, list(S_epi_Seb_vs_all_log_phase_GSEA$ID,
                       S_epi_Seb_vs_all_OS_GSEA$ID,
                       S_epi_Seb_vs_all_stat_phase_GSEA$ID))


S_epi_seb_vs_in_vitro_NESplot <- plot_facetted_NES(input_list=tibble::lst(S_epi_Seb_vs_all_log_phase_GSEA,
                                         S_epi_Seb_vs_all_OS_GSEA, 
                                         S_epi_Seb_vs_all_stat_phase_GSEA),
                  shared_gene_sets = Seb_vs_in_vitro_shared_gene_sets,
                  plot_title="S. epidermidis transcripts, Sebaceous vs in vitro conditions")



ggsave(plot=S_epi_seb_vs_in_vitro_NESplot, filename = "../plots/S_epi_Seb_vs_invitro_NES_plot.pdf", height = 10, width=20)


S_epi_seb_vs_in_vitro_NESplot

```


Enriched gene sets in Staphylococcus epidermidis transcriptomes for toe webs versus in vitro conditions.

Barplots of normalized enrichment scores from gene set enrichment analysis between toe web sites (n=12) and various in vitro conditions (n=9, 6, 9
for log phase, osmotic stress & stationary phase respectively).


```{r}

#no shared gene sets
Tw_vs_in_vitro_shared_gene_sets <- Reduce(intersect, list(S_epi_Tw_vs_all_log_phase_GSEA$ID,
                       S_epi_Tw_vs_all_OS_GSEA$ID,
                       S_epi_Tw_vs_all_stat_phase_GSEA$ID))



#metal ion transport clusters to plot (based on Tw vs log phase DESeq results)

metal_ion_clusters_to_plot <- c("cluster_2534","cluster_2202","cluster_2868","cluster_2072","cluster_13153","cluster_1949","cluster_4306","cluster_21984","cluster_505","cluster_993","cluster_3367","cluster_10128","cluster_2714","cluster_11964","cluster_4685","cluster_14606","cluster_763")


Tw_vs_all_log_phase_l2fc_metal_ion_clust <- Tw_vs_all_log_phase_DE_res  %>% 
  dplyr::filter(pangene %in% metal_ion_clusters_to_plot)

Tw_vs_all_log_phase_l2fc_metal_ion_clust$pangene <- factor(Tw_vs_all_log_phase_l2fc_metal_ion_clust$pangene)

Tw_vs_all_log_phase_l2fc_metal_ion_clust <- Tw_vs_all_log_phase_l2fc_metal_ion_clust %>%
  dplyr::mutate(pangene=fct_reorder(pangene,Description,.fun=identity))


Tw_vs_all_log_phase_l2fc_metal_ion_clust_barplot <- ggplot(Tw_vs_all_log_phase_l2fc_metal_ion_clust, aes(x=pangene,y=log2FoldChange,
                                                     fill=Description)) + 
  geom_col() + 
  scale_fill_manual(values = c23) +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=1))

ggsave(plot=Tw_vs_all_log_phase_l2fc_metal_ion_clust_barplot, 
       filename = "../plots/Tw_vs_all_log_phase_l2fc_metal_ion_clust_barplot.pdf")

Tw_vs_all_log_phase_l2fc_metal_ion_clust_barplot


# copper ion binding clusters to plot (based on Tw_vs_osmotic_stress DESeq results)
copper_ion_binding_clusters_to_plot <- c("cluster_2534","cluster_2072","cluster_13153")

Tw_vs_all_OS_l2fc_Cu_ion_binding <- Tw_vs_all_OS_DE_res  %>% 
  dplyr::filter(pangene %in% copper_ion_binding_clusters_to_plot)

Tw_vs_all_OS_l2fc_Cu_ion_binding <- Tw_vs_all_OS_l2fc_Cu_ion_binding %>%
  dplyr::mutate(pangene=fct_reorder(pangene,Description,.fun=identity))

Tw_vs_all_OS_l2fc_Cu_ion_binding_barplot <- ggplot(Tw_vs_all_OS_l2fc_Cu_ion_binding, 
                                                    aes(x=pangene,y=log2FoldChange,
                                                     fill=Description)) + 
  geom_col() + 
  scale_fill_manual(values = c23) +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=1))


ggsave(plot=Tw_vs_all_OS_l2fc_Cu_ion_binding_barplot, 
       filename = "../plots/Tw_vs_all_OS_l2fc_Cu_ion_binding_barplot.pdf")


Tw_vs_all_OS_l2fc_Cu_ion_binding_barplot

```

Barplots showing log2 fold change of genes related to metal ion transport (top) or
copper ion binding (bottom), relative to in vitro conditions. Libraries representing toe webs had 200,000 S. epidermidis reads each.

# Supplementary figure PCA plots

PCA plot of in vivo vs in vitro samples (in-house samples only)

```{r}


#For In house results only

#batch_correct = FALSE here because the design matrix shouldn't include the batch effect for purposes of PCA.
S_epi_in_vivo_vs_in_vitro_in_house_DE <- run_DESeq_with_inputs_species_specific(input_list = all_S_epi_mtx_counts, 
                                                 lib_vec = get_lib_vec_fn(in_vivo_site = c("Sebaceous",
                                                                                           "Tw"),
                                                                          in_vitro_condition=c("stationary_phase_liquid","osmotic_stress_liquid", "log_phase_liquid"), in_vitro_study = c("this_study")), 
                                                 batch_correct = FALSE, run_test = FALSE)


#PCA plot, no batch correction since this is in vivo vs in vitro data of the same batch
invivo_invitro_in_house_PCA <- plotPCA(S_epi_in_vivo_vs_in_vitro_in_house_DE[["vsd"]], intgroup = "condition", returnData=FALSE) +
  scale_color_manual(values=c("dodgerblue2",  "#E31A1C","green4","black","#6A3D9A"))



ggsave(filename="../plots/invivo_invitro_in_house_PCA.pdf", plot=invivo_invitro_in_house_PCA)


invivo_invitro_in_house_PCA





```


