---
title: "transcriptional_activity_species"
author: "Chiamh"
date: '2024-04-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}
library(tidyverse)
library(ggpubr)
library(reshape2)
library(ggbreak)
```

Load metadata and supporting files.

```{r}

combined_taxa_count_activity_df <- read_tsv("../data/combined_taxa_count_activity_df.tsv",
                                            show_col_types = FALSE)


##For each species, only keep the regions (skin sites) with at least 5 libraries where the species can be detected (metagenomically present)
subsetted_taxa_count_df <- combined_taxa_count_activity_df %>% 
  dplyr::filter(k2_taxon %in% c("Cutibacterium_acnes","Malassezia_globosa",
                                "Malassezia_restricta","Staphylococcus_capitis",
                                "Staphylococcus_epidermidis","Staphylococcus_hominis",
                                "Corynebacterium_tuberculostearicum",
                                "Corynebacterium_kefirresidentii"))

#These species are prevalent across multple skin sites (metagenomic level): Cutibacterium acnes, Malassezia globosa, Malassezia restricta, Staphylococcus capitis, Staphylococcus epidermidis and staphylococcus hominis
taxa_to_KW_test <- c("Cutibacterium_acnes","Malassezia_globosa",
                                "Malassezia_restricta","Staphylococcus_capitis",
                                "Staphylococcus_epidermidis","Staphylococcus_hominis",
                      "Corynebacterium_tuberculostearicum",
                     "Corynebacterium_kefirresidentii")




```


# Transcriptional activity of selected prevalent bacteria across skin sites.

The distribution of the Kruskal-Wallis test statistic approximates a chi-square distribution, with k-1 degrees of freedom, if the number of observations in each group is 5 or more. 

Every taxa tested here is represented in at least 5 or more libraries per site, at the metagenomic level

```{r}

#kruskal wallis test
#return output as a list of lists
subsetted_taxa_KW_test_results <- lapply(taxa_to_KW_test, function(species_name){
  
  species_specific_df <- subsetted_taxa_count_df %>% dplyr::filter(k2_taxon == species_name)
  
  KW_test_result <- kruskal.test(transcriptional_activity ~ region, data = species_specific_df)
  
  #unadjusted for multiple comparisons
  KW_p_value <- KW_test_result$p.value
  

  #For some reason, ggpubr::compare_means does not output identical p values as pairwise.wilcox.test
  #pairwise_wilcox_result <- compare_means(transcriptional_activity ~ region, 
  #                     p.adjust.method = "fdr", method='wilcox.test', data = species_specific_df)
  

  pairwise_wilcox_result_triangular <- pairwise.wilcox.test(species_specific_df$transcriptional_activity,
                                                            species_specific_df$region, 
                                                            p.adjust.method = "fdr", exact=FALSE)
  
  pairwise_wilcox_result_triangular <- melt(pairwise_wilcox_result_triangular$p.value, na.rm=TRUE) %>%
    dplyr::rename(group1=Var1, group2=Var2, p.adj=value)
   
  #add dummy rows for self comparisons
   self_comparison_df <- data.frame(group1=c("Sc","Ch","Ac","Vf","Tw"),
                                    group2=c("Sc","Ch","Ac","Vf","Tw"),
                                    p.adj=NA,
                                    p_adj.signif=NA,
                                    p.adj_exp=NA)
   
  
 pairwise_wilcox_result_triangular$p_adj.signif <- case_when(
  pairwise_wilcox_result_triangular$p.adj >= 0.05 ~ "ns",
  pairwise_wilcox_result_triangular$p.adj < 1e-04  ~ "****",
  pairwise_wilcox_result_triangular$p.adj < 1e-03 ~ "***",
  pairwise_wilcox_result_triangular$p.adj < 1e-02 ~ "**",
  pairwise_wilcox_result_triangular$p.adj < 0.05 ~ "*")
  
  pairwise_wilcox_result_triangular$p_adj.signif <- factor(pairwise_wilcox_result_triangular$p_adj.signif,
                                                  levels=c("****","***","**","*","ns"))

  pairwise_wilcox_result_triangular$p.adj_exp <- format(pairwise_wilcox_result_triangular$p.adj, 
                                                        scientific=TRUE, digits=3)
  
  
  #This is fine because groups 1 and 2 are interchangable for pairwise comparisons
  pairwise_wilcox_result_triangular_mirror <- pairwise_wilcox_result_triangular %>% dplyr::rename(group1=group2,
                                                                                        group2=group1)
  
  
  pairwise_wilcox_result_full <- do.call("rbind",list(pairwise_wilcox_result_triangular, 
                                                pairwise_wilcox_result_triangular_mirror,
                                                self_comparison_df))
  
  pairwise_wilcox_result_full$group1 <- factor(pairwise_wilcox_result_full$group1, levels=c("Sc",
                                                                                  "Ch",
                                                                                  "Ac",
                                                                                  "Vf",
                                                                                  "Tw"))
  
  pairwise_wilcox_result_full$group2 <- factor(pairwise_wilcox_result_full$group2, levels=c("Sc",
                                                                                  "Ch",
                                                                                  "Ac",
                                                                                  "Vf",
                                                                                  "Tw"))
  
  KW_summary_df <- data.frame(k2_taxon=species_name,kruskal_wallis_p_value=KW_p_value)
  
  output <- lst(KW_summary_df, KW_test_result, pairwise_wilcox_result_full)
  
  return(output)
  
  
})


names(subsetted_taxa_KW_test_results) <- taxa_to_KW_test


subsetted_taxa_KW_test_summary <- lapply(taxa_to_KW_test, function(x){
  
  df_out <- subsetted_taxa_KW_test_results[[x]]$KW_summary_df
  
}) %>% do.call("rbind",.)

#adjust for multiple testing

subsetted_taxa_KW_test_summary$kruskal_wallis_p_adj <- p.adjust(subsetted_taxa_KW_test_summary$kruskal_wallis_p_value,
                                                                method="fdr")


```

For box plots, we transform the transcriptional activities by log2(transcription activity + 1).

This transformation preserves the zero values. The Wilcoxon test should give identical results since rank ordering is preserved).

```{r}

#Boxplots for transcriptional activity, per species
#only show the significant differences

subsetted_taxa_count_df$log2_transcriptional_activity <- log2(subsetted_taxa_count_df$transcriptional_activity + 1)

subsetted_taxa_count_df$region <- factor(subsetted_taxa_count_df$region, levels=c("Sc",
                                                                                  "Ch",
                                                                                  "Ac",
                                                                                  "Vf",
                                                                                  "Tw"))
```

For Cutibacterium acnes

```{r}
#########
#C_acnes#
#########

#Displaying the p value in the boxplot is too messy. We can display them in a companion heatmap


C_acnes_trans_act_boxplot <- ggplot(subsetted_taxa_count_df %>% 
         dplyr::filter(k2_taxon == "Cutibacterium_acnes"), 
       aes(x=region,y=log2_transcriptional_activity)) + geom_boxplot() +
  theme_classic() + 
  theme(axis.title.x=element_blank()) + 
  ggtitle("C.acnes transcriptional activity") + scale_y_break(c(2,4), ticklabels=c(4.0, 4.5)) + 
  ylab("log2 (transcriptional activity + 1)")


C_acnes_trans_act_boxplot


C_acnes_pairwise_wilcox_result <- subsetted_taxa_KW_test_results$Cutibacterium_acnes$pairwise_wilcox_result_full



C_acnes_trans_act_p_val_plot <- ggplot(data=C_acnes_pairwise_wilcox_result, aes(x=group1, y = group2, fill=p_adj.signif)) + 
  geom_tile(colour ="white") + geom_text(aes(label=p.adj_exp)) +
  scale_fill_manual(values=c(#"#b62020",
                             "#fe2e2e",
                             "#fe8181",
                             "#ffb6c1",
                             "white")) + 
  scale_x_discrete(expand = c(0, 0)) + 
  scale_y_discrete(expand = c(0, 0)) + 
  theme(axis.ticks = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=1)) +
  theme(panel.background = element_rect(fill = 'white')) +
  theme(axis.title = element_blank()) + ggtitle("C_acnes")




C_acnes_trans_act_p_val_plot





```

For Staphylococcus capitis

```{r}
S_capitis_trans_act_boxplot <- ggplot(subsetted_taxa_count_df %>% 
         dplyr::filter(k2_taxon == "Staphylococcus_capitis"), 
       aes(x=region,y=log2_transcriptional_activity)) + geom_boxplot() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.title.x=element_blank()) + 
  ggtitle("S_capitis transcriptional activity") + ylab("log2 (transcriptional activity + 1)")


S_capitis_trans_act_boxplot


S_capitis_pairwise_wilcox_result <- subsetted_taxa_KW_test_results$Staphylococcus_capitis$pairwise_wilcox_result_full


S_capitis_trans_act_p_val_plot <- ggplot(data=S_capitis_pairwise_wilcox_result, aes(x=group1, y = group2, fill=p_adj.signif)) + 
  geom_tile(colour ="white") + geom_text(aes(label=p.adj_exp)) +
  scale_fill_manual(values=c(#"#b62020",
                             "#fe2e2e",
                             "#fe8181",
                             "#ffb6c1",
                             "white")) + 
  scale_x_discrete(expand = c(0, 0)) + 
  scale_y_discrete(expand = c(0, 0)) + 
  theme(axis.ticks = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=1)) +
  theme(panel.background = element_rect(fill = 'white')) +
  theme(axis.title = element_blank()) + ggtitle("S_capitis")


S_capitis_trans_act_p_val_plot

```
For Staphylococcus epidermidis

```{r}

S_epidermidis_trans_act_boxplot <- ggplot(subsetted_taxa_count_df %>% 
         dplyr::filter(k2_taxon == "Staphylococcus_epidermidis"), 
       aes(x=region,y=log2_transcriptional_activity)) + geom_boxplot() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.title.x=element_blank()) + 
  ggtitle("S_epidermidis transcriptional activity") + ylab("log2 (transcriptional activity + 1)")


S_epidermidis_trans_act_boxplot


S_epidermidis_pairwise_wilcox_result <- subsetted_taxa_KW_test_results$Staphylococcus_epidermidis$pairwise_wilcox_result_full


S_epidermidis_trans_act_p_val_plot <- ggplot(data=S_epidermidis_pairwise_wilcox_result, aes(x=group1, y = group2, fill=p_adj.signif)) + 
  geom_tile(colour ="white") + geom_text(aes(label=p.adj_exp)) +
  scale_fill_manual(values=c("#b62020",
                             "#fe2e2e",
                             "#fe8181",
                             #"#ffb6c1",
                             "white")) + 
  scale_x_discrete(expand = c(0, 0)) + 
  scale_y_discrete(expand = c(0, 0)) + 
  theme(axis.ticks = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=1)) +
  theme(panel.background = element_rect(fill = 'white')) +
  theme(axis.title = element_blank()) + ggtitle("S_epidermidis")


S_epidermidis_trans_act_p_val_plot

```
Assemble subplots into facets

```{r}


bacteria_species_transcriptional_activity <- ggarrange(C_acnes_trans_act_boxplot, S_capitis_trans_act_boxplot, S_epidermidis_trans_act_boxplot,
                                                       C_acnes_trans_act_p_val_plot, S_capitis_trans_act_p_val_plot, S_epidermidis_trans_act_p_val_plot,
                                                       ncol = 3, nrow = 3)

ggsave(plot=bacteria_species_transcriptional_activity, filename="../plots/bacteria_species_transcriptional_activity.pdf",
       width = 12, height = 8)


bacteria_species_transcriptional_activity 

```
## Transcriptional activity of  prevalent fungi across skin sites.

The distribution of the Kruskal-Wallis test statistic approximates a chi-square distribution, with k-1 degrees of freedom, if the number of observations in each group is 5 or more. 

Every taxa tested here is represented in at least 5 or more libraries per site, at the metagenomic level.

```{r}


#########
#M_globosa#
#########


M_globosa_trans_act_boxplot <- ggplot(subsetted_taxa_count_df %>% 
         dplyr::filter(k2_taxon == "Malassezia_globosa"), 
       aes(x=region,y=log2_transcriptional_activity)) + geom_boxplot() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.title.x=element_blank()) + 
  ggtitle("M_globosa transcriptional activity") + ylab("log2 (transcriptional activity + 1)")


M_globosa_trans_act_boxplot


M_globosa_pairwise_wilcox_result <- subsetted_taxa_KW_test_results$Malassezia_globosa$pairwise_wilcox_result_full



M_globosa_trans_act_p_val_plot <- ggplot(data=M_globosa_pairwise_wilcox_result, aes(x=group1, y = group2, fill=p_adj.signif)) + 
  geom_tile(colour ="white") + geom_text(aes(label=p.adj_exp)) +
  scale_fill_manual(values=c(#"#b62020",
                             "#fe2e2e",
                             "#fe8181",
                             #"#ffb6c1",
                             "white")) + 
  scale_x_discrete(expand = c(0, 0)) + 
  scale_y_discrete(expand = c(0, 0)) + 
  theme(axis.ticks = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=1)) +
  theme(panel.background = element_rect(fill = 'white')) +
  theme(axis.title = element_blank()) + ggtitle("M_globosa")



M_globosa_trans_act_p_val_plot



```

For Malassezia restricta

```{r}

#########
#M_restricta#
#########


M_restricta_trans_act_boxplot <- ggplot(subsetted_taxa_count_df %>% 
         dplyr::filter(k2_taxon == "Malassezia_restricta"), 
       aes(x=region,y=log2_transcriptional_activity)) + geom_boxplot() +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        axis.title.x=element_blank()) + 
  ggtitle("M_restricta transcriptional activity") + ylab("log2 (transcriptional activity + 1)")


M_restricta_trans_act_boxplot


M_restricta_pairwise_wilcox_result <- subsetted_taxa_KW_test_results$Malassezia_restricta$pairwise_wilcox_result_full


M_restricta_trans_act_p_val_plot <- ggplot(data=M_restricta_pairwise_wilcox_result, aes(x=group1, y = group2, fill=p_adj.signif)) + 
  geom_tile(colour ="white") + geom_text(aes(label=p.adj_exp)) +
  scale_fill_manual(values=c("#b62020",
                             "#fe2e2e",
                             "#fe8181",
                             "#ffb6c1",
                             "white")) + 
  scale_x_discrete(expand = c(0, 0)) + 
  scale_y_discrete(expand = c(0, 0)) + 
  theme(axis.ticks = element_blank()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.2, hjust=1)) +
  theme(panel.background = element_rect(fill = 'white')) +
  theme(axis.title = element_blank()) + ggtitle("M_restricta")



M_restricta_trans_act_p_val_plot



```


Assemble subplots into facets

```{r}


fungal_species_transcriptional_activity <- ggarrange(M_globosa_trans_act_boxplot, M_restricta_trans_act_boxplot,
                                                       M_globosa_trans_act_p_val_plot, M_restricta_trans_act_p_val_plot,
                                                       ncol = 2, nrow = 2)

ggsave(plot=fungal_species_transcriptional_activity, filename="../plots/fungal_species_transcriptional_activity.pdf",
       width = 12, height = 8)


fungal_species_transcriptional_activity 

```